---
title: "Multivariate Confirmation Analysis - Non-college grads"
editor: visual
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(broom)
library(forcats)
library(stringr)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states)

master <- master.original %>%
  left_join(master.states, by = "PersonID")

kable(head(master))

kable(names(master))
```

# Multivariate Confirmation Analysis: Non-College Completers in Northwest Minnesota

The primary purpose of this supplemental analysis is to examine workforce participation pathways among individuals who did not complete a postsecondary credential. Earlier sections of this analysis demonstrated that among the full Northwest Minnesota population, postsecondary attainment — particularly where and how a credential was earned — was the dominant mechanism associated with sustained regional employment. This section shifts attention to the population for whom that mechanism does not operate.

Among non-college completers, workforce participation pathways cannot be explained by credential depth, institutional sector, or graduation location. Instead, the descriptive and bivariate analyses conducted earlier identified high school career alignment (CTE), post-secondary attendance without completion, community type, and gender as the central differentiators of workforce outcomes in this population.

To strengthen confidence in those findings, a limited multivariate analysis is conducted using logistic regression. As in the full-population analysis, the goal is not causal inference or prediction, but confirmatory testing — specifically, whether the strongest bivariate relationships remain directionally consistent when key predictors are considered simultaneously.

Logistic regression is used here as a confirmatory tool rather than the primary analytic framework.

<br>

# Methodology

## Study Population

The analytic sample is restricted to individuals who graduated from a high school in Northwest Minnesota and did not complete a postsecondary credential. This restriction isolates workforce dynamics among high school graduates entering the labor market directly, with or without some post-secondary exposure.

## Outcome Variable

The primary model predicts:

- **Meaningful employment — region** (binary outcome)

The dependent variable is coded as:

- 1 = Meaningful employment in the Northwest Minnesota region
- 0 = All other workforce outcomes (including meaningful employment elsewhere in Minnesota, not meaningful employment, and no Minnesota employment record)

This specification maintains consistency with earlier modeling while isolating mechanisms specific to the non-college completer population.

## Independent Variables

Because postsecondary completion mechanisms are not applicable in this subgroup, the model focuses on high school preparation, structural background, and local labor market context identified as important in earlier analyses. Notably, the earlier bivariate analysis found that academic readiness measures (MCA scores) predicted out-of-region mobility rather than regional attachment, while CTE engagement was the most consistent and durable predictor of regional workforce attachment across all time points.

Predictors include:

- **Demographic characteristics**
  - `Gender` — the strongest demographic predictor, with males consistently more likely to achieve regional meaningful employment
  - `RaceEthnicity` — included for equity considerations despite negligible overall association
  - `economic.status` — included as a structural disadvantage indicator
  - `SpecialEdStatus` — strongest demographic predictor at year 1 in the bivariate analysis
- **High school and local context**
  - `Dem_Desc` — the strongest high school characteristics predictor, with entirely urban non-completers showing dramatically higher disconnection rates
  - `edr` - regional differences.
  - `avg.unemp.rate` — included with the explicit caveat that its counterintuitive direction likely reflects the North Dakota cross-border employment effect
  - `avg.wages.pct.state` — included as a complementary county-level economic context measure
- **High school accomplishments**
  - `cte.achievement` — the strongest and most consistent predictor of regional workforce attachment identified in the bivariate analysis
  - `MCA.R` — included as the primary academic readiness measure; expected to predict out-of-region rather than regional employment
  - `pseo.participant` — included to test whether early college exposure retains independent influence once other preparation variables are controlled
- **Post-secondary**
  - `attended.ps` — even without completion, post-secondary attendance showed meaningful associations with workforce outcomes

Variables are selected based on:

1.  Strength of association in prior bivariate and persistence analyses
2.  Conceptual plausibility within a non-college pathway
3.  Policy relevance
4.  Avoidance of redundancy

Postsecondary completion variables are excluded by design.

<br>

# Analysis Approach

This multivariate analysis mirrors the structured confirmatory framework used in the full-population models but is adapted to the non-college completer subgroup and the specific findings of the Northwest Minnesota analysis.

## Time-Based Modeling Strategy

Workforce attachment evolves over time. To capture this dynamic process, models are estimated separately at:

- **1 year after high school**
- **5 years after high school**
- **10 years after high school**

This staged approach allows examination of how mechanisms shift across early, mid-term, and longer-term workforce trajectories — a distinction that proved meaningful in the bivariate analysis, where several variables (particularly economic status, special education status, and PSEO participation) showed strong early associations that faded substantially over time.

### Year 1: Immediate Labor Market Entry

The 1-year model captures the initial transition into the labor market. At this stage, high school preparation and career alignment are expected to exert the strongest influence. CTE exposure emerged as the strongest high school-level predictor at year 1 in the bivariate analysis, while post-secondary attendance had its largest association at this time point due to ongoing enrollment. The year 1 model tests whether these early preparation effects hold when structural and contextual factors are considered simultaneously.

### Year 5: Stabilization Phase

The 5-year model reflects early career consolidation. The bivariate analysis showed that CTE's influence stabilized at this point while other variables — particularly economic status and PSEO participation — had largely attenuated. This model evaluates whether CTE retains its independent regional retention effect and whether gender and community type become more prominent as careers develop.

### Year 10: Long-Term Embeddedness

The 10-year model captures longer-term workforce integration. The bivariate analysis suggested that MCA scores became more clearly associated with out-of-region employment by year 10, while CTE remained stable. This model tests whether local labor market context — county wages and unemployment — becomes increasingly dominant over time, and whether the North Dakota border effect continues to complicate interpretation of the unemployment rate variable.

## Modeling Strategy

Within each time period, predictors are introduced in a structured sequence:

### Model 1: Demographic Baseline

Includes demographic characteristics and community type:

- `RaceEthnicity`
- `economic.status`
- `SpecialEdStatus`
- `Gender`

This baseline model evaluates whether structural background independently differentiate regional workforce attachment prior to the inclusion of high school preparation variables. It also establishes whether the gender and urban/rural patterns identified in the bivariate analysis persist in a multivariate context.

### Model 2: Add high school characteristics and  Local Labor Market Context

Adds high school location and county-level economic context to Model 1:

- `Dem_Desc`
- `edr`
- `avg.wages.pct.state`
- `avg.unemp.rate`

### Model 3: Add High School Accomplishments

Adds the following to Model 2:

- `cte.achievement`
- `MCA.R`
- `pseo.participant`

### Model 4: Add post-secondary attendance

Adds the following to Model 3:

- `attended.ps`

This is the full confirmatory specification.

## Evaluation Criteria

Results will be reported using:

- Odds ratios
- 95% confidence intervals
- Statistical significance levels
- Directional interpretation

Model comparison metrics (AIC and pseudo R²) will be used to assess relative improvement across specifications, but emphasis will remain on coefficient magnitude and stability across time points and model specifications.

## Interpretation Framework

Interpretation will focus on:

- Whether CTE retains independent influence across all three time points — the central confirmatory question given its dominance in the bivariate analysis
- Whether gender and community type effects persist once preparation and economic context are controlled
- Whether the unemployment rate's counterintuitive direction is sustained in the multivariate context, providing further confirmation of the North Dakota border effect
- How mechanisms shift from Year 1 to Year 10, and whether structural or contextual factors become increasingly dominant over time

Findings will be described as associations holding other included variables constant. No causal claims will be made.

<br>

## Conceptual Focus

For non-college completers in Northwest Minnesota, this time-structured modeling framework evaluates three evolving mechanisms:

1.  **Career alignment at labor market entry** — captured primarily through CTE engagement and post-secondary attendance
2.  **Structural advantage or disadvantage shaping early stability** — captured through demographic characteristics and community type
3.  **Local economic opportunity structures influencing long-term embeddedness** — captured through county-level wage and unemployment measures, with explicit attention to the North Dakota border effect

Together, these models clarify whether sustained regional employment among non-college completers in Northwest Minnesota reflects early career alignment, structural inequality, local economic context, or a combination of all three — and whether the relative importance of these mechanisms shifts across the first decade after high school.

<br>

# Prep data

The first thing I need to do is setup my dataset for modeling. I'm going to create the workforce participation into a binary and will use grad.year.1, grad.year.5 and grad.year.10 as my modeling year.

```{r}
master.model.1 <- master %>%
  filter(ps.grad == "No") %>%
  select(PersonID, Gender, RaceEthnicity, economic.status, SpecialEdStatus, Dem_Desc, edr, avg.wages.pct.state, avg.unemp.rate, pseo.participant, cte.achievement, MCA.R, attended.ps, grad.year.1) %>%
  filter(grad.year.1 != "After 2023",
         grad.year.1 != "Attending ps",
         RaceEthnicity != "Unknown",
         !is.na(MCA.R)) %>%
  mutate(grad.year.1 = ifelse(grad.year.1 %in% c("Meaningful emp region", "Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", grad.year.1),
         grad.year.1 = str_replace(grad.year.1, ", not attending ps", ""),
         grad.year.1 = ifelse(grad.year.1 == "Meaningful emp region", 1, 0),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         Gender = fct_relevel(Gender, "M"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         attended.ps = fct_relevel(attended.ps, "No"),
         pseo.participant = ifelse(is.na(pseo.participant), 0, pseo.participant),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         economic.status = as.character(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         edr = fct_relevel(edr, "EDR 1 - Northwest"))

kable(head(master.model.1))

kable(names(master.model.1))


master.model.5 <- master %>%
  filter(ps.grad == "No") %>%
  select(PersonID, Gender, RaceEthnicity, economic.status, SpecialEdStatus, edr, Dem_Desc, avg.wages.pct.state, avg.unemp.rate, pseo.participant, cte.achievement, MCA.R, attended.ps, grad.year.5) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         RaceEthnicity != "Unknown",
         !is.na(MCA.R)) %>%
  mutate(grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp region", "Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", grad.year.5),
         grad.year.5 = str_replace(grad.year.5, ", not attending ps", ""),
         grad.year.5 = ifelse(grad.year.5 == "Meaningful emp region", 1, 0),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         Gender = fct_relevel(Gender, "M"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         attended.ps = fct_relevel(attended.ps, "No"),
         pseo.participant = ifelse(is.na(pseo.participant), 0, pseo.participant),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         economic.status = as.character(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         edr = fct_relevel(edr, "EDR 1 - Northwest"))


kable(head(master.model.5))

kable(names(master.model.5))

master.model.10 <- master %>%
  filter(ps.grad == "No",
         RaceEthnicity != "Unknown",
         grad.year.10 != "Attending ps",
         !is.na(MCA.R)) %>%
  select(PersonID, Gender, RaceEthnicity, economic.status, SpecialEdStatus, edr, Dem_Desc, avg.wages.pct.state, avg.unemp.rate, pseo.participant, cte.achievement, MCA.R, attended.ps, grad.year.10) %>%
  filter(grad.year.10 != "After 2023",
         grad.year.10 != "Attending ps") %>%
  mutate(grad.year.10 = ifelse(grad.year.10 %in% c("Meaningful emp region", "Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", grad.year.10),
         grad.year.10 = str_replace(grad.year.10, ", not attending ps", ""),
         grad.year.10 = ifelse(grad.year.10 == "Meaningful emp region", 1, 0),
         RaceEthnicity = fct_relevel(RaceEthnicity, "White"),
         Gender = fct_relevel(Gender, "M"),
         cte.achievement = fct_relevel(cte.achievement, "No CTE"),
         attended.ps = fct_relevel(attended.ps, "No"),
         pseo.participant = ifelse(is.na(pseo.participant), 0, pseo.participant),
         pseo.participant = as.character(pseo.participant),
         pseo.participant = fct_relevel(pseo.participant, "0"),
         economic.status = as.character(economic.status),
         economic.status = fct_relevel(economic.status, "0"),
         SpecialEdStatus = as.character(SpecialEdStatus),
         SpecialEdStatus = fct_relevel(SpecialEdStatus, "0"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
         edr = fct_relevel(edr, "EDR 1 - Northwest"))


kable(head(master.model.10))

kable(names(master.model.10))

```

We now have `r comma(nrow(master.model.5))` individuals to model with `r ncol(master.model.5) - 2` variables not including `grad.year.5`.

<br>

# Modeling - `grad.year.1`

## Model 1 — Year 1: Demographic Baseline

**Overview** Model 1 establishes the demographic baseline for predicting meaningful regional employment at one year post-graduation among non-college completers (N=18,983). The model includes race/ethnicity, economic status, special education status, and gender. Variance inflation factors were all near 1.0, indicating no multicollinearity concerns. The Hosmer-Lemeshow test indicated some misfit at the baseline stage (χ²=22.99, df=6, p\<.001), which is expected given the limited explanatory variables included. The model explains a very small proportion of variance in the outcome (Nagelkerke R²=0.005), confirming that demographic characteristics alone account for little of the variation in regional workforce attachment at this early stage.

- Race/ethnicity: AI non-completers show no significant association relative to White non-completers (OR=0.951, 95% CI \[0.851, 1.061\], p=.371)
- Race/ethnicity: Asian/PI non-completers show significantly lower odds relative to White non-completers (OR=0.519, 95% CI \[0.337, 0.769\], p=.002), the largest disparity in the model
- Race/ethnicity: Black non-completers show significantly lower odds relative to White non-completers (OR=0.730, 95% CI \[0.560, 0.938\], p=.016)
- Race/ethnicity: Hispanic non-completers show significantly lower odds relative to White non-completers (OR=0.824, 95% CI \[0.689, 0.980\], p=.031)
- Economic status: no significant association with regional employment at year 1 (OR=0.990, 95% CI \[0.923, 1.061\], p=.772), consistent with the negligible bivariate finding for this subgroup
- Special education status: associated with significantly lower odds of regional employment (OR=0.855, 95% CI \[0.790, 0.925\], p\<.001)
- Gender: females show significantly lower odds of regional employment than males (OR=0.839, 95% CI \[0.783, 0.898\], p\<.001), consistent with the male-dominated industry composition identified in the bivariate analysis

**Summary** The demographic baseline reveals modest but significant disparities in regional employment among non-college completers at year 1. Asian/PI non-completers show the largest disparity of any group, followed by Black and Hispanic non-completers, while AI non-completers show no significant association — a different pattern from the full-population and postsecondary completer analyses, where AI disparities were consistently among the largest. Special education status and gender both show significant, moderate negative associations, aligning with the bivariate findings that identified these as among the strongest early demographic differentiators. Economic status shows no significant association, consistent with the bivariate analysis. The very small Nagelkerke R² (0.005) confirms that demographic characteristics alone explain very little of the variation in regional workforce attachment for non-completers at year 1, with high school characteristics, local labor market context, and especially high school accomplishments (given CTE's dominance in the bivariate analysis) expected to add substantially more explanatory power in subsequent blocks.

```{r}
model1 <- glm(
  grad.year.1 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender,
  family = binomial,
  data = master.model.1
)

model1

# Odds ratios with 95% confidence intervals
exp(cbind(OR = coef(model1), confint(model1)))

# Wald test p-values and full summary
summary(model1)

# Nagelkerke R²
pscl::pR2(model1)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model1$y, fitted(model1))


# VIF values
car::vif(model1)

or_plot <- tidy(model1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "Gender" = "Gender: ",
      "SpecialEdStatus" = "Special Ed: ",
      "Dem_Desc" = "Dem_Desc: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2 — Year 1: Adding High School Characteristics

**Overview** Model 2 adds Dem_Desc, edr, average wages as a percent of state, and average unemployment rate to the demographic baseline, maintaining the same sample of 18,983 non-college completers. VIF values remained well below 2.0 for all predictors (highest GVIF\^(1/(2\*Df))=1.38 for avg.wages.pct.state), indicating no multicollinearity concerns. Model fit improved meaningfully over the demographic baseline (ΔAIC=-154.36, Nagelkerke R²=0.018, up from 0.005), and the Hosmer-Lemeshow test improved to marginal acceptability (χ²=14.697, df=8, p=.065).

**Stability of Demographic Coefficients** Demographic coefficients from Model 1 show mixed stability with the addition of high school characteristics and local labor market context. - Gender remains virtually unchanged (OR=0.833, p\<.001) - AI non-completers remain non-significant (OR=0.947, p=.367) - Asian/PI non-completers remain stable, the largest disparity in the model (OR=0.510, p=.001) - Black non-completers attenuate to marginal non-significance (OR from 0.730 to 0.801, p=.094), a notable change from Model 1 - Hispanic non-completers attenuate to non-significance (OR from 0.824 to 0.888, p=.192), also a notable change from Model 1 - Economic status remains non-significant (OR=0.959, p=.240) - Special education status remains virtually unchanged (OR=0.853, p\<.001)

**Dem_Desc** Rural classification of the high school district shows the strongest association of any variable in Model 2, consistent with its identification as the strongest high school characteristics predictor in the bivariate analysis. - Entirely urban non-completers show significantly lower odds of regional employment relative to entirely rural non-completers (OR=0.593, 95% CI \[0.497, 0.707\], p\<.001), confirming the "urban penalty" identified in the bivariate analysis - Town/rural mix non-completers show significantly higher odds relative to entirely rural non-completers (OR=1.253, 95% CI \[1.112, 1.415\], p\<.001) - Urban/town/rural mix non-completers also show significantly higher odds (OR=1.200, 95% CI \[1.050, 1.373\], p=.008)

**EDR** No significant association with regional employment. - EDR 2 - Headwaters shows no significant association relative to EDR 1 - Northwest (OR=1.010, p=.882) - EDR 4 - West Central shows no significant association relative to EDR 1 - Northwest (OR=1.030, p=.509)

**Average Wages (% of State)** No significant association with regional employment (OR=1.091, 95% CI \[0.648, 1.832\], p=.742).

**Average Unemployment Rate** No significant association with regional employment (OR=1.002, 95% CI \[0.978, 1.025\], p=.896). This directly contradicts the counterintuitive positive bivariate relationship for this variable, indicating that the bivariate pattern — interpreted as a North Dakota cross-border employment effect — does not hold once demographic characteristics and community type are controlled simultaneously.

**Summary** The addition of Dem_Desc, edr, and local labor market context produces a meaningful improvement in model fit over the demographic baseline (ΔAIC=-154.36, ΔNagelkerke R²=+0.013). Dem_Desc shows the strongest association in the model, confirming the urban penalty and rural/mixed advantage identified in the bivariate analysis. Both EDR and average unemployment rate show no significant independent association, consistent with EDR's identification as one of the weakest bivariate predictors and providing multivariate disconfirmation of the unemployment rate's counterintuitive bivariate direction — suggesting that pattern reflected confounding with community type or demographic composition rather than an independent local labor market mechanism. The attenuation of Black and Hispanic disparities to non-significance indicates that some of their Model 1 disadvantage was explained by differences in community type and geographic distribution rather than persisting independently.

```{r}
model2 <- glm(
  grad.year.1 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.1
)

model2

nobs(model1)
nobs(model2)

AIC(model1)
AIC(model2)

exp(cbind(OR = coef(model2), confint(model2)))

# Wald test p-values and full summary
summary(model2)

# Nagelkerke R²
pscl::pR2(model2)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model1$y, fitted(model2))

# VIF values
car::vif(model2)


or_plot <- tidy(model2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 3 — Year 1: Adding High School Accomplishments

**Overview** Model 3 adds cte.achievement, MCA.R, and pseo.participant to the previous model, maintaining the same sample of 18,983 non-college completers. VIF values remained well below 2.0 for all predictors (highest GVIF\^(1/(2\*Df))=1.40 for avg.wages.pct.state), indicating no multicollinearity concerns. Model fit improved meaningfully over Model 2 (ΔAIC=-162.33, Nagelkerke R²=0.031, up from 0.018), and the Hosmer-Lemeshow test indicates good model fit (χ²=6.337, df=8, p=.610), the best-fitting model in this sequence so far.

**Stability and Attenuation of Previous Coefficients** The addition of high school accomplishment variables produces meaningful changes in several coefficients from Model 2. - Gender attenuates further (OR from 0.833 to 0.892), remaining significant (p=.001) - AI non-completers remain non-significant (OR from 0.947 to 0.981, p=.755) - Asian/PI non-completers remain stable, the largest disparity in the model (OR=0.510, p=.001) - Black non-completers remain non-significant (OR from 0.801 to 0.826, p=.153) - Hispanic non-completers remain non-significant (OR from 0.888 to 0.904, p=.271) - Economic status remains non-significant (OR from 0.959 to 0.942, p=.102) - Special education status strengthens (OR from 0.853 to 0.806), remaining highly significant (p\<.001) - Dem_Desc associations remain stable and significant across all categories - EDR remains non-significant for both categories - Average wages as a percent of state becomes significant (OR from 1.091, p=.742 to 1.720, 95% CI \[1.015, 2.908\], p=.043), a notable reversal from Model 2 - Average unemployment rate remains non-significant (OR=1.003, p=.823), continuing to disconfirm the bivariate counterintuitive pattern

**CTE Achievement** CTE engagement shows the strongest associations of any variable in Model 3, confirming its identification as the strongest and most consistent predictor in the bivariate analysis. - CTE Concentrators or Completors show significantly higher odds of regional employment relative to No CTE non-completers (OR=1.712, 95% CI \[1.553, 1.888\], p\<.001) - CTE Participants show a smaller but still substantial positive association (OR=1.338, 95% CI \[1.202, 1.490\], p\<.001) - A clear dose-response pattern is evident, consistent with the bivariate finding

**MCA Reading** No significant association with regional employment at year 1 (OR=0.979, 95% CI \[0.941, 1.019\], p=.297). This is a notable attenuation from the bivariate relationship — Reading showed the strongest effect size among the three MCA subjects at year 1 in the bivariate analysis, but once CTE engagement and PSEO participation are controlled, it retains no independent association, suggesting its bivariate relationship was substantially confounded with career-pathway variables rather than reflecting an independent academic-readiness effect at this early stage.

**PSEO Participation** A significant negative association with regional employment (OR=0.773, 95% CI \[0.694, 0.859\], p\<.001), consistent with PSEO functioning as a marker of postsecondary-oriented pathway selection even among students who ultimately do not complete a credential.

**Summary** The addition of high school accomplishment variables produces a meaningful improvement in model fit (ΔAIC=-162.33, ΔNagelkerke R²=+0.013) and the best-fitting model in the Year 1 sequence. CTE achievement dominates this block, confirming its bivariate identification as the central mechanism associated with regional workforce attachment among non-completers. Neither MCA Reading nor the local labor market variables from Model 2 show consistently significant independent effects once CTE and PSEO are controlled — Reading's null result is a meaningful departure from its bivariate strength, indicating that academic proficiency's apparent relationship with regional employment operates largely through its correlation with career-pathway engagement rather than independently. PSEO participation shows a significant negative association, consistent with early disconnection from CTE-aligned regional pathways. The continued attenuation of racial and ethnic disparities to non-significance, alongside stable special education and gender effects, suggests that whatever demographic gaps exist at year 1 for non-completers are substantially — though not entirely — explained by community type and career preparation differences.

```{r}
model3 <- glm(
  grad.year.1 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    MCA.R +
    pseo.participant,
  family = binomial,
  data = master.model.1
)

model3

nobs(model3)

AIC(model1, model2, model3)

exp(cbind(OR = coef(model3), confint(model3)))

# Wald test p-values and full summary
summary(model3)

# Nagelkerke R²
pscl::pR2(model3)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model3$y, fitted(model3))

# VIF values
car::vif(model3)

or_plot <- tidy(model3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4 — Year 1: Adding Post-secondary attendance

**Overview** Model 4 adds attended.ps to the full Model 3 specification, maintaining the same sample of 18,983 non-college completers. VIF values remained well below 2.0 for all predictors including attended.ps (GVIF\^(1/(2\*Df))=1.02), indicating no multicollinearity concerns. The addition of this variable produces the largest single improvement in model fit observed across any block in the Year 1 sequence (ΔAIC=-189.70, Nagelkerke R²=0.046, up from 0.031), confirming attended.ps as the dominant predictor at this early stage identified in the bivariate analysis. The Hosmer-Lemeshow test indicates some misfit (χ²=21.858, df=8, p=.005), likely attributable to sample size sensitivity given the pattern observed in other analyses in this project.

**Stability and Attenuation of Previous Coefficients** The addition of postsecondary attendance produces meaningful changes in several coefficients from Model 3. - Gender attenuates further (OR from 0.892 to 0.869), remaining significant (p\<.001) - Black non-completers move toward marginal significance (OR from 0.826 to 0.789, p=.077), a notable strengthening - Asian/PI non-completers remain stable, the largest racial/ethnic disparity in the model (OR=0.499, p=.001) - AI and Hispanic non-completers remain non-significant - Economic status remains non-significant (OR=0.962, p=.295) - Special education status strengthens further (OR from 0.806 to 0.836), remaining highly significant (p\<.001) - Dem_Desc associations strengthen modestly and remain highly significant across all categories - EDR 2 - Headwaters moves to marginal significance (OR from 1.124, p=.102 to 1.128, p=.092), still short of conventional significance - EDR 4 - West Central attenuates to complete non-significance (OR from 1.050 to 1.003, p=.941) - Average wages as a percent of state strengthens further (OR from 1.720 to 1.923), remaining significant (p=.015) - Average unemployment rate remains non-significant and reverses to a small negative direction (OR=0.994, p=.658) - CTE Concentrators/Completors and Participants remain stable and highly significant, with modest attenuation (OR from 1.712 to 1.695; OR from 1.338 to 1.316) - MCA Reading moves to marginal significance in the expected negative direction (OR from 0.979, p=.297 to 0.964, p=.075) - PSEO participation remains stable and significant (OR from 0.773 to 0.781, p\<.001)

**Postsecondary Attendance** A strong, significant positive association with regional employment at year 1 (OR=1.692, 95% CI \[1.571, 1.821\], p\<.001), the largest single association introduced in this model and the strongest overall confirmation of the bivariate finding that even unsuccessful postsecondary attendance is strongly tied to early regional workforce attachment — likely reflecting continued local ties and delayed full labor market entry rather than a causal preparation effect.

**Summary** The addition of postsecondary attendance produces the largest improvement in model fit of any block in the Year 1 sequence (ΔAIC=-189.70, ΔNagelkerke R²=+0.015), confirming its bivariate identification as the strongest early differentiator among non-college completers. This is the full confirmatory specification for Year 1: CTE achievement and postsecondary attendance emerge as the two dominant, independent associates of regional employment, with Dem_Desc's urban penalty persisting robustly across all four models. MCA Reading's shift toward the expected negative direction once attended.ps is separated out, though still short of significance, suggests some of its bivariate relationship was previously entangled with postsecondary-attendance status. The persistence of special education status and gender disparities, alongside the continued attenuation of most racial/ethnic disparities to non-significance, indicates that community type, career alignment, and postsecondary exposure jointly explain much — though not all — of the demographic variation in regional employment identified at baseline.

```{r}
model4 <- glm(
  grad.year.1 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    MCA.R +
    pseo.participant +
    attended.ps,
  family = binomial,
  data = master.model.1
)

model4

nobs(model4)

AIC(model1, model2, model3, model4)

exp(cbind(OR = coef(model4), confint(model4)))

# Wald test p-values and full summary
summary(model4)

# Nagelkerke R²
pscl::pR2(model4)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model4$y, fitted(model4))

# VIF values
car::vif(model4)

or_plot <- tidy(model4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

# Modeling - `grad.year.5`

## Model 1 — Year 5: Demographic Baseline

**Overview** Model 1 establishes the demographic baseline for predicting meaningful regional employment at five years post-graduation among non-college completers (N=19,626). The model includes race/ethnicity, economic status, special education status, and gender. Variance inflation factors were all near 1.0, indicating no multicollinearity concerns. The Hosmer-Lemeshow test indicates good model fit (χ²=8.198, df=6, p=.224). The model explains a very small proportion of variance in the outcome (Nagelkerke R²=0.009), similar in magnitude to the year 1 baseline, confirming that demographic characteristics alone continue to explain little of the variation in regional workforce attachment.

- Race/ethnicity: AI non-completers show no significant association relative to White non-completers (OR=0.942, 95% CI \[0.837, 1.059\], p=.321), unchanged from year 1
- Race/ethnicity: Asian/PI non-completers show significantly lower odds relative to White non-completers (OR=0.565, 95% CI \[0.383, 0.810\], p=.003)
- Race/ethnicity: Black non-completers show significantly lower odds relative to White non-completers (OR=0.491, 95% CI \[0.369, 0.641\], p\<.001), a substantially larger disparity than at year 1 (OR=0.730)
- Race/ethnicity: Hispanic non-completers show significantly lower odds relative to White non-completers (OR=0.613, 95% CI \[0.506, 0.739\], p\<.001), also a considerably larger disparity than at year 1 (OR=0.824)
- Economic status: no significant association with regional employment at year 5 (OR=1.040, 95% CI \[0.974, 1.111\], p=.240), consistent with the bivariate finding that its influence fades after year 1
- Special education status: no significant association at year 5 (OR=0.986, 95% CI \[0.914, 1.065\], p=.727), a complete reversal from its significant year 1 association (OR=0.855), confirming the bivariate finding that special education status is the strongest demographic predictor only at year 1
- Gender: females show significantly lower odds of regional employment than males (OR=0.770, 95% CI \[0.721, 0.821\], p\<.001), a widening gap compared to year 1 (OR=0.839)

**Summary** The demographic baseline at year 5 shows a markedly different pattern from year 1. Special education status and economic status, both significant early differentiators, fade to complete non-significance by year 5, exactly as the bivariate analysis predicted. In their place, racial and ethnic disparities — particularly for Black and Hispanic non-completers — and the gender gap both strengthen substantially, suggesting that early transitional disadvantages give way to more structural, persistent disparities as careers develop. The very small Nagelkerke R² (0.009) confirms that demographic characteristics alone continue to explain little of the variation in regional workforce attachment, with high school characteristics, local labor market context, and CTE engagement expected to add more explanatory power in subsequent blocks, as they did at year 1.

```{r}
model1 <- glm(
  grad.year.5 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender,
  family = binomial,
  data = master.model.5
)

model1

# Odds ratios with 95% confidence intervals
exp(cbind(OR = coef(model1), confint(model1)))

# Wald test p-values and full summary
summary(model1)

# Nagelkerke R²
pscl::pR2(model1)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model1$y, fitted(model1))


# VIF values
car::vif(model1)

or_plot <- tidy(model1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "Gender" = "Gender: ",
      "SpecialEdStatus" = "Special Ed: ",
      "Dem_Desc" = "Dem_Desc: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2 — Year 5: Adding High School Characteristics

**Overview** Model 2 adds Dem_Desc, edr, average wages as a percent of state, and average unemployment rate to the demographic baseline, maintaining the same sample of 19,626 non-college completers. VIF values remained well below 2.0 for all predictors (highest GVIF\^(1/(2\*Df))=1.43 for avg.wages.pct.state), indicating no multicollinearity concerns. Model fit improved meaningfully over the demographic baseline (ΔAIC=-251.37, Nagelkerke R²=0.029, up from 0.009), though the Hosmer-Lemeshow test indicates some misfit (χ²=18.874, df=8, p=.016).

**Stability of Demographic Coefficients** Demographic coefficients from Model 1 show mixed stability with the addition of high school characteristics and local labor market context. - Gender remains virtually unchanged (OR=0.762, p\<.001) - AI non-completers move toward marginal significance (OR from 0.942 to 0.895, p=.082), a change from year 1 where AI showed no association at any stage - Asian/PI non-completers remain stable (OR=0.571, p=.003) - Black non-completers remain stable and highly significant, the largest disparity in the model (OR=0.521, p\<.001) - Hispanic non-completers remain stable and significant (OR=0.676, p\<.001) - Economic status remains non-significant (OR=0.995, p=.885) - Special education status remains non-significant (OR=0.986, p=.721)

**Dem_Desc** Rural classification of the high school district shows the strongest association of any variable in Model 2, consistent with the pattern at year 1. - Entirely urban non-completers show significantly lower odds of regional employment relative to entirely rural non-completers (OR=0.460, 95% CI \[0.389, 0.543\], p\<.001), a slightly larger urban penalty than at year 1 (OR=0.593) - Town/rural mix non-completers show a smaller but still significant positive association (OR=1.137, 95% CI \[1.020, 1.270\], p=.021), a weaker association than at year 1 (OR=1.253) - Urban/town/rural mix non-completers show no significant association (OR=1.033, p=.613), a change from the significant positive association at year 1

**EDR** A modest association emerges at year 5, unlike the complete non-significance observed at year 1. - EDR 2 - Headwaters shows no significant association relative to EDR 1 - Northwest (OR=0.976, p=.723) - EDR 4 - West Central shows a significant positive association relative to EDR 1 - Northwest (OR=1.102, 95% CI \[1.014, 1.197\], p=.022), a new finding compared to year 1

**Average Wages (% of State)** No significant association with regional employment (OR=0.727, 95% CI \[0.449, 1.175\], p=.194).

**Average Unemployment Rate** A significant positive association with regional employment (OR=1.031, 95% CI \[1.008, 1.054\], p=.008), a notable reversal from the complete non-significance observed at year 1. This strengthens the case for the North Dakota cross-border employment interpretation described in the bivariate analysis, suggesting the pattern becomes a stable, independent association only after the immediate post-graduation transition period.

**Summary** The addition of Dem_Desc, edr, and local labor market context produces a meaningful improvement in model fit over the demographic baseline (ΔAIC=-251.37, ΔNagelkerke R²=+0.019). Dem_Desc continues to show the strongest association in the model, with a slightly deepened urban penalty compared to year 1. Unlike at year 1, average unemployment rate now shows a significant positive association, and EDR 4 - West Central emerges as significant — both new findings at this stage that were absent at year 1, suggesting geographic and local economic context become more consequential as careers progress from immediate post-graduation transition into longer-term attachment.

```{r}
model2 <- glm(
  grad.year.5 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.5
)

model2


nobs(model1)
nobs(model2)

AIC(model1)
AIC(model2)

exp(cbind(OR = coef(model2), confint(model2)))

# Wald test p-values and full summary
summary(model2)

# Nagelkerke R²
pscl::pR2(model2)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model1$y, fitted(model2))

# VIF values
car::vif(model2)

or_plot <- tidy(model2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 3 — Year 5: Adding High School Accomplishments

**Overview** Model 3 adds cte.achievement, MCA.R, and pseo.participant to the previous model, maintaining the same sample of 19,626 non-college completers. VIF values remained well below 2.0 for all predictors (highest GVIF\^(1/(2\*Df))=1.46 for avg.wages.pct.state), indicating no multicollinearity concerns. Model fit improved meaningfully over Model 2 (ΔAIC=-141.13, Nagelkerke R²=0.039, up from 0.029), and the Hosmer-Lemeshow test indicates good model fit (χ²=7.281, df=8, p=.507), the best-fitting model in the year 5 sequence so far.

**Stability and Attenuation of Previous Coefficients** The addition of high school accomplishment variables produces meaningful changes in several coefficients from Model 2. - Gender attenuates modestly (OR from 0.762 to 0.803), remaining highly significant (p\<.001) - AI non-completers attenuate to non-significance (OR from 0.895 to 0.918, p=.183) - Asian/PI non-completers remain stable (OR=0.578, p=.004) - Black non-completers remain stable and highly significant, the largest disparity in the model (OR=0.529, p\<.001) - Hispanic non-completers remain stable and significant (OR=0.678, p\<.001) - Economic status remains non-significant (OR=0.975, p=.465) - Special education status re-emerges as significant (OR from 0.986, p=.721 to 0.902, p=.020), an unusual non-monotonic pattern — significant at baseline, non-significant once geography and labor market context are added, then significant again once academic preparation variables are controlled - Dem_Desc associations remain stable and significant for entirely urban and town/rural mix categories - EDR 4 - West Central remains significant and stable (OR=1.115, p=.011) - EDR 2 - Headwaters remains non-significant (OR=1.102, p=.159) - Average wages as a percent of state remains non-significant (OR=1.218, p=.431) - Average unemployment rate remains significant and stable (OR=1.029, p=.013)

**CTE Achievement** CTE engagement continues to show large, significant positive associations, though somewhat attenuated from year 1. - CTE Concentrators or Completors show significantly higher odds of regional employment relative to No CTE non-completers (OR=1.595, 95% CI \[1.458, 1.745\], p\<.001), a modest attenuation from year 1 (OR=1.712) but still the strongest association in the model - CTE Participants show a smaller but still substantial positive association (OR=1.380, 95% CI \[1.252, 1.521\], p\<.001), slightly strengthening compared to year 1 (OR=1.338)

**MCA Reading** A significant negative association with regional employment (OR=0.941, 95% CI \[0.906, 0.977\], p=.002), the expected direction and the first point in this analysis where Reading shows an independent effect — unlike year 1, where it was completely non-significant. This confirms the bivariate finding that academic proficiency's association with out-of-region mobility strengthens as careers develop.

**PSEO Participation** A significant negative association with regional employment (OR=0.851, 95% CI \[0.769, 0.940\], p=.002), a smaller magnitude than at year 1 (OR=0.773) but remaining significant.

**Summary** The addition of high school accomplishment variables produces a meaningful improvement in model fit (ΔAIC=-141.13, ΔNagelkerke R²=+0.011). CTE achievement remains the dominant predictor in this block, though somewhat attenuated from year 1. MCA Reading's emergence as a significant independent predictor at year 5 — where it showed no effect at year 1 — confirms the bivariate pattern that academic readiness measures increasingly point toward out-of-region mobility as time passes. The re-emergence of special education status as significant, after being fully explained by geography and labor market context in Model 2, suggests its disadvantage operates through a different mechanism than initially apparent — one connected to academic preparation and career pathway variables rather than purely structural or geographic factors.

```{r}
model3 <- glm(
  grad.year.5 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    MCA.R +
    pseo.participant,
  family = binomial,
  data = master.model.5
)

model3

nobs(model3)

AIC(model1, model2, model3)

exp(cbind(OR = coef(model3), confint(model3)))

# Wald test p-values and full summary
summary(model3)

# Nagelkerke R²
pscl::pR2(model3)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model3$y, fitted(model3))

# VIF values
car::vif(model3)

or_plot <- tidy(model3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4 — Year 5: Adding Post-secondary

**Overview** Model 4 adds attended.ps to the full Model 3 specification, maintaining the same sample of 19,626 non-college completers. VIF values remained well below 2.0 for all predictors including attended.ps (GVIF\^(1/(2\*Df))=1.03), indicating no multicollinearity concerns. The addition of this variable produces essentially no improvement in model fit (ΔAIC=-0.92, Nagelkerke R²=0.040, unchanged from Model 3), a striking contrast with year 1, where attended.ps produced the largest single block improvement in the entire sequence (ΔAIC=-189.70). This confirms the bivariate finding that postsecondary attendance's association with regional employment is a short-term, mechanical effect that has almost entirely dissipated by year 5.

**Stability of Previous Coefficients** All coefficients from Model 3 remain essentially unchanged with the addition of attended.ps, consistent with its negligible independent contribution at this stage. CTE achievement, MCA Reading, PSEO participation, Dem_Desc, EDR 4 - West Central, and average unemployment rate all retain their Model 3 magnitudes and significance levels within rounding.

**Postsecondary Attendance** A marginal, not quite statistically significant association with regional employment at year 5 (OR=1.059, 95% CI \[0.992, 1.130\], p=.087), a dramatic decline from its dominant year 1 association (OR=1.692, p\<.001). This confirms the bivariate finding that postsecondary attendance's influence is largely mechanical — reflecting ongoing enrollment status shortly after high school — rather than a durable marker of workforce trajectory.

**Summary** The addition of postsecondary attendance produces no meaningful improvement in model fit at year 5, a striking reversal from its role as the dominant predictor at year 1. This is the full confirmatory specification for year 5: CTE achievement remains the strongest and most stable predictor across both time points examined so far, while MCA Reading has emerged as a significant independent predictor pointing toward out-of-region mobility — a relationship entirely absent at year 1. Dem_Desc's urban penalty persists robustly, and average unemployment rate's positive association, absent at year 1, is now significant, reinforcing the North Dakota cross-border employment interpretation as a pattern that solidifies over time rather than appearing immediately after graduation. The near-total disappearance of attended.ps's effect confirms that its year 1 dominance reflected transitional enrollment status rather than any lasting relationship with regional workforce attachment.

```{r}
model4 <- glm(
  grad.year.5 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    MCA.R +
    pseo.participant +
    attended.ps,
  family = binomial,
  data = master.model.5
)

model4

nobs(model4)

AIC(model1, model2, model3, model4)

exp(cbind(OR = coef(model4), confint(model4)))

# Wald test p-values and full summary
summary(model4)

# Nagelkerke R²
pscl::pR2(model4)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model4$y, fitted(model4))

# VIF values
car::vif(model4)

or_plot <- tidy(model4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

# Modeling - `grad.year.10`

## Model 1 — Year 10: Demographic Baseline

**Overview** Model 1 establishes the demographic baseline for predicting meaningful regional employment at ten years post-graduation among non-college completers (N=11,448). The model includes race/ethnicity, economic status, special education status, and gender. Variance inflation factors were all near 1.0, indicating no multicollinearity concerns. The Hosmer-Lemeshow test indicates some misfit (χ²=20.697, df=5, p\<.001), consistent with the pattern observed at year 1. The model explains a very small proportion of variance in the outcome (Nagelkerke R²=0.009), similar in magnitude to years 1 and 5, confirming that demographic characteristics alone continue to explain little of the variation in regional workforce attachment even at this longer-term stage.

- Race/ethnicity: AI non-completers show no significant association relative to White non-completers (OR=1.014, 95% CI \[0.863, 1.188\], p=.864), stable with the null findings at years 1 and 5
- Race/ethnicity: Asian/PI non-completers show significantly lower odds relative to White non-completers (OR=0.370, 95% CI \[0.198, 0.637\], p\<.001), the largest racial/ethnic disparity observed at any time point in this analysis
- Race/ethnicity: Black non-completers show significantly lower odds relative to White non-completers (OR=0.517, 95% CI \[0.337, 0.764\], p=.001), similar in magnitude to year 5 (OR=0.491)
- Race/ethnicity: Hispanic non-completers show significantly lower odds relative to White non-completers (OR=0.734, 95% CI \[0.562, 0.947\], p=.020), a smaller disparity than at year 5 (OR=0.613)
- Economic status: no significant association with regional employment at year 10 (OR=1.004, 95% CI \[0.918, 1.097\], p=.936), consistent with years 1 and 5
- Special education status: a significant positive association with regional employment (OR=1.120, 95% CI \[1.009, 1.241\], p=.032) — a complete reversal from the significant negative association at year 1 (OR=0.855) and the non-significant association at year 5 (OR=0.986)
- Gender: females show significantly lower odds of regional employment than males (OR=0.768, 95% CI \[0.704, 0.838\], p\<.001), stable with the pattern at years 1 and 5

**Summary** The demographic baseline at year 10 reveals the most striking trajectory shift in this analysis: special education status moves from a significant early disadvantage to a significant long-term advantage, suggesting that whatever early barriers special education non-completers face in entering the regional workforce reverse over the following decade, with this group ultimately showing stronger regional attachment than their peers. The Asian/PI disparity, meanwhile, grows to its largest magnitude of any time point examined. Gender and Black disparities remain stable and persistent across all three time points, while economic status continues to show no association at any stage. The very small Nagelkerke R² (0.009), essentially unchanged from years 1 and 5, confirms that demographic characteristics alone remain weak explanators of regional attachment throughout the first decade after high school, with CTE engagement and geographic context expected to continue driving most of the explanatory power, as at earlier time points.

```{r}
model1 <- glm(
  grad.year.10 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender,
  family = binomial,
  data = master.model.10
)

model1

# Odds ratios with 95% confidence intervals
exp(cbind(OR = coef(model1), confint(model1)))

# Wald test p-values and full summary
summary(model1)

# Nagelkerke R²
pscl::pR2(model1)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model1$y, fitted(model1))


# VIF values
car::vif(model1)

or_plot <- tidy(model1, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "Gender" = "Gender: ",
      "SpecialEdStatus" = "Special Ed: ",
      "Dem_Desc" = "Dem_Desc: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )



```

<br>

## Model 2 — Year 10: Adding High School Characteristics

**Overview** Model 2 adds Dem_Desc, edr, average wages as a percent of state, and average unemployment rate to the demographic baseline, maintaining the same sample of 11,448 non-college completers. VIF values remained well below 2.0 for all predictors (highest GVIF\^(1/(2\*Df))=1.48 for avg.wages.pct.state), indicating no multicollinearity concerns. Model fit improved substantially over the demographic baseline (ΔAIC=-204.02, Nagelkerke R²=0.037, up from 0.009), though the Hosmer-Lemeshow test indicates continued misfit (χ²=25.243, df=8, p=.001).

**Stability of Demographic Coefficients** Demographic coefficients from Model 1 show mixed stability with the addition of high school characteristics and local labor market context. - Gender remains virtually unchanged (OR=0.759, p\<.001) - AI non-completers remain non-significant (OR=0.947, p=.527) - Asian/PI non-completers remain the largest disparity in the model and strengthen modestly (OR=0.359, p=.001) - Black non-completers remain stable and significant (OR=0.527, p=.002) - Hispanic non-completers attenuate to non-significance (OR from 0.734 to 0.809, p=.115), a change from year 1 and year 5, where Hispanic disparities were consistently significant - Economic status remains non-significant (OR=0.947, p=.236) - Special education status remains significant and positive (OR=1.132, p=.020), consistent with the reversal observed in Model 1

**Dem_Desc** Rural classification of the high school district shows its strongest association of any time point examined. - Entirely urban non-completers show significantly lower odds of regional employment relative to entirely rural non-completers (OR=0.405, 95% CI \[0.320, 0.511\], p\<.001), the deepest urban penalty observed at any time point (compared to OR=0.593 at year 1 and OR=0.460 at year 5) - Town/rural mix and urban/town/rural mix non-completers show no significant association, a change from earlier time points

**EDR** A significant association emerges for the first time in this analysis. - EDR 2 - Headwaters shows a significant negative association relative to EDR 1 - Northwest (OR=0.793, 95% CI \[0.659, 0.955\], p=.014), the first significant EDR finding at any time point in the non-completer analysis - EDR 4 - West Central shows no significant association (OR=0.944, p=.305)

**Average Wages (% of State)** A marginal, not quite statistically significant negative association with regional employment (OR=0.580, 95% CI \[0.304, 1.104\], p=.098).

**Average Unemployment Rate** A significant positive association with regional employment (OR=1.070, 95% CI \[1.035, 1.107\], p\<.001), continuing to strengthen from year 1 (non-significant) through year 5 (OR=1.031) to year 10, confirming the methodology's prediction that local labor market context becomes increasingly dominant over the first decade after high school.

**Summary** The addition of Dem_Desc, edr, and local labor market context produces a substantial improvement in model fit over the demographic baseline (ΔAIC=-204.02, ΔNagelkerke R²=+0.028), the largest Model 2 improvement of any time point examined. Dem_Desc shows its deepest urban penalty at this stage, and EDR emerges as significant for the first time, with Headwaters non-completers showing notably lower odds of regional employment. The monotonic strengthening of average unemployment rate across all three time points provides strong confirmation of the North Dakota cross-border employment interpretation as a long-term, cumulative pattern rather than an immediate one.

```{r}
model2 <- glm(
  grad.year.10 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.model.10
)

model2

model2


nobs(model1)
nobs(model2)

AIC(model1)
AIC(model2)

exp(cbind(OR = coef(model2), confint(model2)))

# Wald test p-values and full summary
summary(model2)

# Nagelkerke R²
pscl::pR2(model2)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model1$y, fitted(model2))

# VIF values
car::vif(model2)

or_plot <- tidy(model2, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: "
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 3 — Year 10: Adding High School Accomplishments

**Overview** Model 3 adds cte.achievement, MCA.R, and pseo.participant to the previous model, maintaining the same sample of 11,448 non-college completers. VIF values remained well below 2.0 for all predictors (highest GVIF\^(1/(2\*Df))=1.54 for avg.wages.pct.state), indicating no multicollinearity concerns. Model fit improved meaningfully over Model 2 (ΔAIC=-95.82, Nagelkerke R²=0.050, up from 0.037), and the Hosmer-Lemeshow test indicates good model fit (χ²=8.586, df=8, p=.378), the best-fitting model in the year 10 sequence.

**Stability and Attenuation of Previous Coefficients** The addition of high school accomplishment variables produces meaningful changes in several coefficients from Model 2. - Gender attenuates modestly (OR from 0.759 to 0.792), remaining highly significant (p\<.001) - AI non-completers remain non-significant (OR=0.940, p=.479) - Asian/PI non-completers remain the largest disparity in the model (OR=0.353, p\<.001) - Black non-completers remain stable and significant (OR=0.516, p=.002) - Hispanic non-completers move toward marginal significance (OR from 0.809 to 0.795, p=.091) - Economic status moves toward marginal significance (OR from 0.947 to 0.915, p=.057), a change from complete non-significance at years 1 and 5 - Special education status reverses again, attenuating to non-significance (OR from 1.132, p=.020 to 0.942, p=.325) — the third distinct pattern this variable has shown within the year 10 sequence alone - Dem_Desc's entirely urban penalty remains strong and significant (OR=0.472) - EDR 2 - Headwaters attenuates to non-significance (OR from 0.793, p=.014 to 0.866, p=.133), indicating its Model 2 association was partially explained by CTE and academic preparation composition - Average unemployment rate strengthens further (OR from 1.070 to 1.110), remaining highly significant (p\<.001) - Average wages as a percent of state remains non-significant (OR=1.307, p=.435)

**CTE Achievement** CTE engagement remains significant, though the most attenuated of any time point examined. - CTE Concentrators or Completors show significantly higher odds of regional employment (OR=1.505, 95% CI \[1.331, 1.704\], p\<.001), continuing the gradual attenuation observed across time points (year 1: OR=1.712; year 5: OR=1.595; year 10: OR=1.505) - CTE Participants show a smaller positive association (OR=1.270, 95% CI \[1.109, 1.456\], p\<.001), similarly attenuated from earlier time points

**MCA Reading** A strong, significant negative association with regional employment (OR=0.845, 95% CI \[0.803, 0.889\], p\<.001), the largest MCA Reading effect observed at any time point, continuing its monotonic strengthening from null at year 1 through OR=0.941 at year 5.

**PSEO Participation** No significant association with regional employment (OR=1.020, 95% CI \[0.863, 1.202\], p=.813), a reversal from the significant negative associations observed at both year 1 (OR=0.773) and year 5 (OR=0.851) — PSEO's early disconnection effect appears to fully resolve by year 10.

**Summary** The addition of high school accomplishment variables produces a meaningful improvement in model fit (ΔAIC=-95.82, ΔNagelkerke R²=+0.013). CTE achievement, while showing its most attenuated effect of any time point, remains significant and substantively important. MCA Reading's continued strengthening confirms academic proficiency as an increasingly powerful long-term predictor of out-of-region mobility among non-completers. The resolution of PSEO's early disadvantage and the further instability of special education status suggest that some early-career disconnection patterns are transitional rather than permanent, while CTE's dose-response relationship and Dem_Desc's urban penalty remain the most durable mechanisms across the full ten-year window examined.

```{r}
model3 <- glm(
  grad.year.10 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    MCA.R +
    pseo.participant,
  family = binomial,
  data = master.model.10
)

model3

nobs(model3)

AIC(model1, model2, model3)

exp(cbind(OR = coef(model3), confint(model3)))

# Wald test p-values and full summary
summary(model3)

# Nagelkerke R²
pscl::pR2(model3)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model3$y, fitted(model3))

# VIF values
car::vif(model3)

or_plot <- tidy(model3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

## Model 4 — Year 10: Adding Post-secondary Attendance

**Overview** Model 4 adds attended.ps to the full Model 3 specification, maintaining the same sample of 11,448 non-college completers. VIF values remained well below 2.0 for all predictors including attended.ps (GVIF\^(1/(2\*Df))=1.03), indicating no multicollinearity concerns. The addition of this variable produces negligible improvement in model fit (ΔAIC=-1.42, Nagelkerke R²=0.051, essentially unchanged from Model 3), consistent with the bivariate finding that postsecondary attendance's influence is short-lived rather than durable.

**Stability and Attenuation of Previous Coefficients** Most Model 3 coefficients remain stable with the addition of attended.ps. - Economic status becomes significant for the first time at any stage of this analysis (OR from 0.915, p=.057 to 0.909, 95% CI \[0.830, 0.996\], p=.040), a new finding that emerges only once postsecondary attendance is isolated as its own term - Gender, race/ethnicity, Dem_Desc, EDR, average unemployment rate, CTE achievement, and MCA Reading all remain essentially unchanged from Model 3

**Postsecondary Attendance** A marginal, not quite statistically significant negative association with regional employment (OR=0.920, 95% CI \[0.842, 1.005\], p=.064). This represents a genuine sign reversal, not merely attenuation, from the strong positive association observed at year 1 (OR=1.692, p\<.001) and the smaller positive association at year 5 (OR=1.059, p=.087). By year 10, non-completers who attended postsecondary trend toward lower, not higher, odds of regional employment relative to those who never attended — the opposite of the pattern observed immediately after high school.

**Summary** The addition of postsecondary attendance produces no meaningful improvement in model fit at year 10, confirming the bivariate finding that its influence is concentrated in the immediate post-graduation period. Unlike years 1 and 5, however, its coefficient does not simply fade toward zero — it crosses into negative territory, suggesting that non-completers who attempted but did not finish postsecondary education may face a modest long-term disadvantage in regional employment relative to those who entered the workforce directly, a pattern worth further investigation given it runs counter to the variable's dominant early-career role. This is the full confirmatory specification for year 10: MCA Reading and average unemployment rate have both grown into the strongest independent associates of regional employment by this stage, CTE achievement remains significant though increasingly attenuated, and Dem_Desc's urban penalty persists as the most stable finding across the entire three-time-point analysis.

```{r}
model4 <- glm(
  grad.year.10 ~
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    Gender +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    cte.achievement +
    MCA.R +
    pseo.participant + 
    attended.ps,
  family = binomial,
  data = master.model.10
)

model4

nobs(model4)

AIC(model1, model2, model3, model4)

exp(cbind(OR = coef(model4), confint(model4)))

# Wald test p-values and full summary
summary(model4)

# Nagelkerke R²
pscl::pR2(model4)

# Hosmer-Lemeshow test
ResourceSelection::hoslem.test(model4$y, fitted(model4))

# VIF values
car::vif(model4)

or_plot <- tidy(model4, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Make labels nicer (edit these replacements to match your exact factor labels)
    term = str_replace_all(term, c(
      "RaceEthnicity" = "Race: ",
      "economic.status" = "Economic status: ",
      "SpecialEdStatus" = "Special Ed: ",
      "pseo.participant" = "PSEO: ",
      "cte.achievement" = "CTE: ",
      "MCA.R" = "MCA Reading",
      "MCA.M" = "MCA Math",
      "Dem_Desc" = "Dem_Desc: ",
      "Gender" = "Gender: ",
      "took.ACT" = "Took ACT: ",
      "avg.unemp.rate" = "Avg unemp rate",
      "avg.wages.pct.state" = "Avg wages pct state"
    )),
    # Put most important effects near the top (optional: tweak ordering later)
    term = fct_reorder(term, estimate)
  )

ggplot(or_plot, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = c(0.5, 0.75, 1.25, 1.5, 2),
             color = "grey85") +
  geom_point() +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    orientation = "y",
    width = 0.2
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = c("0.5", "0.75", "1.0", "1.25", "1.5", "2.0")
  ) +
  theme_line +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL
  )


```

<br>

# Multivariate Confirmation Analysis Summary — Non-College Completers in Northwest Minnesota

## Multivariate Confirmation Analysis Summary — Non-College Completers in Northwest Minnesota

**CTE Engagement Is the Most Consistent and Durable Associate**

CTE achievement is the only high school accomplishment variable that shows a significant, stable, positive association with regional employment at all three time points and in every model specification. CTE Concentrators or Completors show odds of regional employment roughly 50-70% higher than No CTE non-completers across the full ten-year window (final model OR=1.695 at year 1, OR=1.591 at year 5, OR=1.507 at year 10), with CTE Participants showing a smaller but similarly durable positive association throughout. While the magnitude attenuates modestly over time, CTE never loses significance and never changes direction — the only variable in this analysis with that distinction. This confirms CTE as the central, policy-actionable mechanism connecting high school preparation to long-term regional workforce retention among non-completers.

**Post-Secondary Attendance and Academic Proficiency Shift Direction Over Time**

Two variables show genuine sign reversals across the ten-year window rather than simple attenuation. Attended post-secondary is the dominant predictor at year 1 (OR=1.692, p\<.001) — the largest single association in the entire year 1 sequence — but weakens to marginal non-significance by year 5 (OR=1.059, p=.087) and crosses into negative territory by year 10 (OR=0.920, p=.064). This is not fading toward null; it is a reversal, suggesting non-completers who attempted postsecondary education without finishing may carry a modest long-term disadvantage relative to those who entered the workforce directly, despite an initial advantage.

MCA Reading shows the opposite trajectory — a monotonic strengthening rather than a reversal. It shows no independent association at year 1 (p=.075, marginal), becomes significant at year 5 (OR=0.938, p=.001), and shows its strongest effect at year 10 (OR=0.850, p\<.001). This confirms academic proficiency as an increasingly powerful predictor of out-of-region mobility the further out from high school graduation one looks — the opposite pattern from attended post-secondary's declining relevance.

**Local Labor Market Context Becomes Increasingly Important Over Time**

Average county unemployment rate shows the clearest monotonic pattern in the entire analysis, strengthening steadily across all three time points and every model specification: non-significant at year 1 (OR=0.994, p=.658), significant at year 5 (OR=1.028, p=.016), and most strongly significant at year 10 (OR=1.102, p\<.001). This provides increasingly strong confirmation of the North Dakota cross-border employment interpretation established in the bivariate analysis — individuals with no Minnesota employment record are concentrated in lower-unemployment border counties, and this pattern becomes more pronounced, not less, as more time passes since graduation. Average wages as a percent of state, by contrast, is significant only at year 1 (OR=1.923, p=.015) and non-significant at years 5 and 10, suggesting county wage levels matter for immediate post-graduation decisions but not for longer-term attachment.

**The Urban Penalty Is Structural, Persistent, and Growing**

Dem_Desc's entirely-urban penalty is the single most stable and deepening finding in this analysis. Entirely urban non-completers show significantly lower odds of regional employment relative to entirely rural non-completers at every time point and every model specification, with the penalty growing steadily more severe: OR=0.628 at year 1, OR=0.487 at year 5, OR=0.470 at year 10 (final model estimates). Unlike nearly every other variable examined, this association shows no meaningful attenuation once high school accomplishments, local labor market context, or postsecondary attendance are controlled — confirming that community type operates as an independent, structural mechanism rather than a proxy for demographic or preparation differences.

**Gender Disparities Are Persistent and Structural**

Females show significantly lower odds of regional employment than males at every time point and every model specification, with the gap widening slightly over time (final model OR=0.869 at year 1, OR=0.802 at year 5, OR=0.795 at year 10). The consistency and modest strengthening of this disparity, net of community type, CTE engagement, academic preparation, and local labor market context, is consistent with the bivariate interpretation that it reflects the male-dominated industry composition of the Northwest Minnesota regional economy rather than a preparation or access gap.

**Racial and Ethnic Disparities Persist Net of All Variables**

Racial and ethnic patterns vary considerably by group and by time point, in contrast to the more uniform demographic patterns above. Asian/PI non-completers show a significant disadvantage at every time point, growing to the largest disparity observed anywhere in this analysis by year 10 (OR=0.352). Black non-completers show a marginal association at year 1 that strengthens to a stable, significant disparity by years 5 and 10 (final model OR≈0.52 at both stages). Hispanic non-completers show a significant disparity at year 5 only, with non-significant associations at years 1 and 10. AI non-completers show no significant association at any time point or model specification — a pattern distinct from the full-population and postsecondary-completer analyses, where AI disparities were consistently among the largest observed. The persistence of Black and Asian/PI disparities net of community type, CTE engagement, academic preparation, and postsecondary attendance indicates structural barriers to regional employment that are not explained by the mechanisms examined in this analysis.

**Other Structural Variables Shift Over Time**

Special education status, economic status, PSEO participation, and EDR each show unstable or time-limited patterns rather than consistent trajectories. Special education status is significantly associated with lower regional employment odds at years 1 and 5 (final model OR=0.836 and OR=0.908 respectively) but becomes non-significant by year 10, and shows notable instability across model specifications within the year 10 sequence specifically — significant positive in the demographic baseline, then non-significant once academic preparation is controlled — suggesting its early disadvantage resolves over time through mechanisms only partially captured by the variables in this model. Economic status shows no significant association at years 1 or 5 but becomes significant at year 10 only once postsecondary attendance is isolated as its own term (OR=0.909, p=.040), a delayed and unexpected finding. PSEO participation shows a significant negative association at years 1 and 5 that fully resolves to non-significance by year 10. EDR shows no significant association at year 1, a significant Headwaters disadvantage at year 5 that persists only through Model 2 before attenuating, and no significant association in the final year 10 model — the weakest and least stable predictor examined, consistent with its identification as one of the two weakest bivariate predictors overall.

**Overall Conclusion**

Across the ten-year window examined, three findings stand out as the most robust and policy-relevant. First, CTE engagement is the only mechanism examined that shows a consistently positive, significant, and durable association with regional employment at every stage — the clearest and most actionable lever for regional workforce retention among non-college completers. Second, community type operates as a persistent and deepening structural factor independent of preparation, academic proficiency, or postsecondary exposure, with entirely urban non-completers facing a growing regional employment penalty over time. Third, the mechanisms shaping regional attachment shift substantially across the decade: early-career outcomes are dominated by postsecondary attendance status and CTE engagement, while longer-term outcomes are increasingly shaped by academic proficiency (in a direction that favors out-of-region mobility) and local labor market context, particularly the North Dakota cross-border employment pattern. This time-varying structure confirms that a single cross-sectional model would have substantially mischaracterized the mechanisms driving regional workforce attachment among non-college completers, and that policy interventions calibrated to the immediate post-graduation period — such as those targeting postsecondary transition support — may need to be paired with distinct, longer-horizon strategies addressing local economic opportunity and CTE-pathway investment to sustain regional retention over the full first decade of the workforce.
