---
title: "Multivariate Confirmation Analysis on Persistent Disconnection"
editor: visual
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  select(5,6) %>%
  unique() %>%
  mutate(edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/Master-non-college-grads-disconnection.csv") 

kable(names(master.original))

kable(head(master.original))
```

# Multivariate Confirmation Analysis: Disconnection Persistence Among Non–Postsecondary Graduates

## Introduction

The bivariate analysis of workforce disconnection among non-college graduates identified two distinct disconnection mechanisms operating within a single combined measure. Non-sustained employment (working, but below the sustained-employment threshold) and absence of any Minnesota employment record behave as structurally different phenomena: the former is front-loaded and largely resolves within five years, while the latter is bimodal and behaves as a durable end-state once established. Separating these two mechanisms revealed that different demographic and high-school-accomplishment variables predict each one — most notably, Special Education status and English learner status appeared to drive no-record persistence specifically, while Economic status and Gender appeared to drive non-sustained persistence specifically, and different racial/ethnic groups showed disparities concentrated in one mechanism or the other.

This multivariate analysis exists to confirm whether that mechanism-split pattern holds once all variables are considered simultaneously, rather than one at a time. This step is necessary because earlier work in this project has shown that bivariate relationships — particularly those involving English learner status — can substantially attenuate or shift once other variables, especially missing-data patterns in academic assessment scores, are accounted for in the same model. Before this mechanism-split finding is presented as a confirmed result, it should be tested under the same conditions.

## Methodology

### Study Population

The analytic sample is restricted to individuals who did **not** complete a postsecondary credential, consistent with the non-college-graduate population used throughout this project. The sample is further restricted to individuals who graduated (or exited postsecondary attendance, for those who attended without completing) after 2010, consistent with the cohort restriction applied elsewhere in this project to avoid the PSEO participation tracking gap present in earlier cohorts.

### Dependent Variables

This analysis models two outcomes, each derived from the disconnection-persistence categories established in the bivariate analysis. Each graduate's workforce status was tracked across the first five years after exiting formal education (high school for direct-to-workforce graduates, or postsecondary attendance for those who attended without completing).

- **Persistent non-sustained employment** — binary (1 = non-sustained workforce participation in 4 or 5 of the first 5 years, 0 = otherwise). This isolates the instability mechanism identified in the bivariate work.
- **Persistent no-record status** — binary (1 = no Minnesota employment record in 4 or 5 of the first 5 years, 0 = otherwise). This isolates the disappearance-from-the-system mechanism identified in the bivariate work.

Each outcome is modeled separately, in its own specification, rather than as levels of a single combined variable. This mirrors the logic of the bivariate analysis: combining the two mechanisms into one model would risk re-obscuring the distinction this analysis is specifically designed to confirm.

### Independent Variables

The following indepdent variables were confirmed associated to our dependent variables in the bivariate analysis. These will all be considered for use in the modeling.

**Demographics**

- `Gender`
- `RaceEthnicity`
- `economic.status`
- `SpecialEdStatus`
- `english.learner`

**High school characteristics**

- `Dem_Desc`
- `edr`
- `avg.unemp.rate`
- `avg.wages.pct.state`

**High school accomplishments**

- `pseo.participant`
- `ap.exam`
- `took.ACT`
- `ACTCompositeScore` (recoded to 0 for non-test-takers, entered alongside `took.ACT`, per the missing-indicator approach established earlier in this project)
- `cte.achievement`
- `total.cte.courses.taken`
- `sat.taken`

**Postsecondary variable**

- `attended.ps` — included specifically because it was the single strongest predictor of overall disconnection in the bivariate analysis, and it is the one variable found to meaningfully affect both disconnection mechanisms rather than only one.

### Modeling Approach

Each outcome (persistent non-sustained status, persistent no-record status) is modeled using a staged block-entry sequence, consistent with the modeling structure used throughout this project's other multivariate confirmation analyses. Predictors are introduced in four stages to make attenuation of any individual variable's effect visible and attributable to a specific block, rather than only observable in a single final specification.

**Model 1: Demographics**

- Demographics: `Gender`, `RaceEthnicity`, `economic.status`, `SpecialEdStatus`, `english.learner`,

**Model 2: add High School Characteristics**

- High school characteristics: `Dem_Desc`, `edr`, `avg.unemp.rate`, `avg.wages.pct.state`

**Model 3: Add High School Accomplishments**

- Adds: `pseo.participant`, `took.ACT`, `ACTCompositeScore.filled`, `ap.exam`, `cte.achievement`, `total.cte.courses.taken`, `sat.taken`

**Model 4: Add Postsecondary Attempt**

- Adds: `attended.ps`

Each model is compared to the prior one using a likelihood ratio test to confirm it adds significant explanatory power, with AIC tracked across the sequence to summarize overall model improvement. This staged structure is applied identically to both outcomes, allowing direct comparison of how each mechanism's predictors evolve across the same three blocks.

### Evaluation Criteria

Results will be reported using odds ratios, 95% confidence intervals, and statistical significance levels. Interpretation will focus specifically on whether the demographic variables identified as mechanism-specific in the bivariate analysis (Special Education and English learner status for no-record persistence; Economic status and Gender for non-sustained persistence; racial/ethnic groups for whichever mechanism their bivariate disparity was concentrated in) remain significant, attenuate, or reverse once the full variable set is included. Findings will be described as associations holding other included variables constant. No causal claims will be made.

<br>

# Prepping the data

The modeling dataset begins from the merged file containing both the disconnection-persistence categories (established in the bivariate analysis) and the full set of demographic, high school, and postsecondary variables used throughout this project. Three preparation steps are applied before modeling.

**Cohort restriction.** The sample is restricted to graduates from 2011 onward (`grad.year > 2010`). This mirrors the cohort restriction applied throughout this project's other analyses of non-college graduates: PSEO participation was not reliably tracked prior to 2011 (missingness runs 92-100% in the 2008-2010 cohorts, dropping to near-zero by 2014), and including those cohorts would risk misclassifying untracked PSEO participants as confirmed non-participants.

**Reference level assignment.** Categorical variables are releveled to match the reference categories used consistently throughout this project (male, White, non-economically-disadvantaged, non-Special-Education, non-English-learner, EDR 5 - North Central, no CTE involvement, non-participant/non-test-taker for the binary accomplishment variables). This keeps odds ratios in this analysis directly comparable in direction and interpretation to every other model built in this project.

**Missing-indicator recodes for two variables with structural or correlated missingness:**

- **ACT composite score.** Students who did not take the ACT have no composite score by definition — confirmed here to be a fully deterministic relationship with `took.ACT` (zero mismatches between the two). `ACTCompositeScore` is recoded to 0 for non-test-takers and entered into the model alongside `took.ACT`, so that `took.ACT` absorbs the effect of test participation itself while the filled score captures the achievement gradient among test-takers only.

**Dependent variable construction.** The two outcomes tested in this analysis are built directly from the bivariate persistence categories, collapsed to binary indicators:

- `persistent.non_sustained` — 1 if the graduate was non-sustained in 4 or 5 of their first 5 years post-education-exit, 0 otherwise.

- `persistent.no_record` — 1 if the graduate had no Minnesota employment record in 4 or 5 of those years, 0 otherwise.

```{r}
final_ivs <- c(
  # Demographics
  "Gender",
  "RaceEthnicity",
  "economic.status",
  "SpecialEdStatus",
  "english.learner",

  # High school characteristics
  "Dem_Desc",
  "edr",
  "avg.unemp.rate",
  "avg.wages.pct.state",
  
  # High school accomplishments
  "pseo.participant",
  "took.ACT",
  "ACTCompositeScore",
  "ap.exam",
  "sat.taken",
  "cte.achievement",
  "total.cte.courses.taken",
  
  # Postsecondary
  "attended.ps"
)

master.disc <- master.original %>%
  filter(grad.year > 2010,
         !is.na(pseo.participant),
         grad.year < 2020,
         RaceEthnicity != "Unknown") %>%
  select(PersonID, all_of(final_ivs), non_sustained.cat, no_record.cat) %>%
  mutate(
    Gender              = fct_relevel(Gender, "M"),
    RaceEthnicity       = fct_relevel(RaceEthnicity, "White"),
    economic.status = as_factor(economic.status),
    economic.status     = fct_relevel(economic.status, "0"),
    SpecialEdStatus = as_factor(SpecialEdStatus),
    SpecialEdStatus     = fct_relevel(SpecialEdStatus, "0"),
    english.learner = as_factor(english.learner),
    english.learner     = fct_relevel(english.learner, "0"),
    Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural"),
    edr = fct_relevel(edr, "EDR 6W- Upper Minnesota Valley"),
    pseo.participant = as_factor(pseo.participant),
    pseo.participant    = fct_relevel(pseo.participant, "0"),
    took.ACT            = fct_relevel(took.ACT, "No"),
    ap.exam = as_factor(ap.exam),
    ap.exam             = fct_relevel(ap.exam, "0"),
    sat.taken = as_factor(sat.taken),
    sat.taken = fct_relevel(sat.taken, "0"),
    cte.achievement     = fct_relevel(cte.achievement, "No CTE"),
    attended.ps         = fct_relevel(attended.ps, "No"),
    ACTCompositeScore.filled = ifelse(is.na(ACTCompositeScore), 0, ACTCompositeScore),
    persistent.non_sustained = ifelse(non_sustained.cat == "Persistently non-sustained", 1, 0),
    persistent.no_record     = ifelse(no_record.cat == "Persistently no-record", 1, 0)
  ) 
nrow(master.disc)

kable(names(master.disc))

kable(head(master.disc))
```

<br>

# Modeling: persistent non-sustained WF

## Model 1: Demographics

This baseline model examines whether demographic characteristics are associated with **persistent non-sustained employment specifically** — the job-instability mechanism isolated in the bivariate decomposition — among non-postsecondary graduates, prior to accounting for high school context or accomplishments. N = 8,707.

Before the individual coefficients: this model provides a direct multivariate test of the mechanism-split hypothesis stated in the introduction to this analysis, and the results are worth checking against that stated hypothesis explicitly. The introduction characterized Economic status and Gender as the bivariate drivers of non-sustained persistence specifically, and Special Education status and English learner status as drivers of no-record persistence specifically. **The results below confirm the Economic status and Gender half of that claim, but do not match the Special Education half** — Special Education status shows a strong, highly significant association with persistent *non-sustained* employment here, consistent with the actual bivariate finding reported earlier in this project (Special Ed status was the second-strongest demographic predictor in the non-sustained-only table, Cramér's V = 0.076, and showed almost no relationship to no-record status, V = 0.037). This is worth flagging as a correction to the introduction's framing rather than an unexpected multivariate result — the bivariate evidence always pointed to Special Ed status as a non-sustained-specific driver, and this model confirms that.

**Special education status** shows the second-largest association in this model: students with special education status have **roughly 46 percent higher odds** of persistent non-sustained employment (OR = 1.456, p \< .001). This directly confirms the bivariate finding and corrects the introduction's stated hypothesis — Special Ed status is a job-instability-specific risk factor, not a no-record-specific one.

**Gender** shows a significant association, confirming the bivariate hypothesis: **women have roughly 47 percent higher odds** of persistent non-sustained employment than men (OR = 1.475, p \< .001) — the largest coefficient in the model. This is consistent with gender's role throughout the bivariate work as a genuine, non-attenuating predictor of job instability specifically.

**Race/ethnicity** shows a single clear, significant association, confirming the bivariate finding precisely. Relative to White graduates, **Black graduates have roughly 82 percent higher odds** of persistent non-sustained employment (OR = 1.822, p \< .001) — directly confirming the bivariate table's finding that Black graduates showed the highest persistent non-sustained rate of any racial/ethnic group (21.5%), a pattern the bivariate no-record table showed was *not* shared with American Indian graduates. **American Indian graduates show a modest, non-significant trend in the same direction** (OR = 1.374, p = .120) — consistent with the bivariate finding that American Indian graduates' elevated overall disconnection risk was concentrated in no-record status, not job instability, so a weaker, non-significant signal here is expected rather than surprising. **Asian/PI and Hispanic graduates are both non-significant** (OR = 1.066, p = .707; OR = 1.061, p = .569), also consistent with their comparatively muted bivariate showings in the non-sustained table.

**Economic status** confirms the bivariate hypothesis directly: economically disadvantaged graduates show **roughly 28 percent higher odds** of persistent non-sustained employment (OR = 1.278, p \< .001) — consistent with economic status's role as the *strongest* demographic predictor in the bivariate non-sustained table (Cramér's V = 0.086), where it had been nearly invisible in the combined measure.

**English learner status shows no association at all** (OR = 1.004, p = .971) — a genuine null, and directly consistent with the bivariate finding that English learner status's signal was concentrated in the no-record table rather than the non-sustained table. This is the one part of the introduction's stated hypothesis that aligns cleanly with both the bivariate evidence and this model.

**Model fit:** AIC = 6,804.56, residual deviance = 6,786.6 on 8,698 df (null deviance = 6,890.5 on 8,706 df), a modest improvement over the null model (pseudo R² ≈ 0.015). VIF values are low across all predictors (max GVIF\^(1/2Df) ≈ 1.20), indicating no multicollinearity concerns at this demographics-only stage.

**Summary:** At the demographics-only stage, this model provides strong multivariate confirmation of the bivariate mechanism-split story for non-sustained employment specifically: Gender, Special Education status, and Black race/ethnicity all emerge as significant, substantial predictors, while economic status confirms its bivariate role as the single strongest demographic driver of job instability. English learner status is correctly null here, consistent with its no-record-specific bivariate signal. The one point requiring a correction to the analysis's own stated framing is Special Education status — the introduction characterized it as a no-record-specific driver, but both the bivariate evidence and this model agree it is, in fact, one of the strongest non-sustained-specific predictors identified in this entire project. This should be corrected in the final write-up rather than carried forward as stated. As high school context and accomplishments are added, it will be worth watching whether Gender, Special Education status, and the Black disparity persist — and, in the companion no-record model, whether Special Education status remains appropriately null there, providing the mirror-image confirmation this analysis is designed to produce.

```{r}
model1_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.disc
)

model1_non_sustained

summary(model1_non_sustained)

nobs(model1_non_sustained)
AIC(model1_non_sustained)
exp(cbind(OR = coef(model1_non_sustained), confint(model1_non_sustained)))
1 - (model1_non_sustained$deviance / model1_non_sustained$null.deviance)

vif(model1_non_sustained)
```

<br>

## Model 2: Add High School Characteristics

This model adds high school characteristics (community type, EDR, local wage and unemployment context) to the demographic block, examining whether these community-level factors are associated with persistent non-sustained employment, and whether they attenuate or strengthen the demographic associations identified in Model 1. N = 8,707.

**Gender, economic status, and special education status all remain essentially unchanged from Model 1**, confirming these three associations are not explained by the community or regional economic context these non-completers come from. **Women continue to show roughly 48 percent higher odds** of persistent non-sustained employment (OR = 1.480, p \< .001). **Special education status remains the second-largest effect**, with roughly 46 percent higher odds (OR = 1.463, p \< .001). **Economic status remains essentially stable**, with roughly 27 percent higher odds (OR = 1.274, p \< .001). All three continue to confirm their bivariate roles as genuine, job-instability-specific predictors.

**Race/ethnicity** remains essentially stable, with the Black disparity softening only slightly. Relative to White graduates, **Black graduates continue to show significantly higher odds** of persistent non-sustained employment, now roughly 72 percent higher (OR = 1.718, p = .001), down modestly from Model 1's 82 percent but still the clearest race/ethnicity finding in the model. **American Indian, Asian/PI, and Hispanic graduates all remain non-significant**, consistent with Model 1 — the Asian/PI point estimate moves somewhat further from the null (OR = 1.171, p = .361) but does not approach significance.

**English learner status remains non-significant** (OR = 1.007, p = .944), consistent with Model 1 and with the bivariate expectation that this variable's signal belongs to the no-record mechanism, not this one.

**Community type (`Dem_Desc`)** shows a newly significant, though modest, association: relative to entirely rural graduates, those from a **town/rural mix show roughly 24 percent higher odds** of persistent non-sustained employment (OR = 1.239, p = .049) — a borderline but real finding. This is a genuinely interesting result to flag against the bivariate work: the bivariate non-sustained table found entirely rural graduates showed the *lowest* persistent non-sustained rate of the three community types (10.7%), with town/rural mix showing the highest (13.6%) — this model confirms that bivariate direction holds even after controlling for demographics. **Urban/town/rural mix remains non-significant** (OR = 1.029, p = .839).

**EDR** produces the most substantively important new finding in this model. **EDR 8–Southwest shows significantly *lower* odds of persistent non-sustained employment relative to EDR 6W** (OR = 0.780, p = .004) — a roughly 22 percent reduction. This directly confirms the bivariate hypothesis raised in the high-school-characteristics writeup: EDR 8 graduates, who showed the *highest* persistent no-record rate of any region in the bivariate no-record table, show here — now confirmed in a multivariate specification — that they are *not* at elevated risk of job instability specifically; if anything, they're somewhat protected against it relative to EDR 6W. **EDR 6E remains non-significant** (OR = 1.080, p = .402), consistent with its comparatively muted bivariate showing in the non-sustained table.

**Local average wages and local unemployment rate are both non-significant** (wages: OR = 1.884, p = .204; unemployment: OR = 0.967, p = .203). This is consistent with the bivariate finding that local wages showed no relationship to non-sustained employment at all, and roughly consistent with local unemployment's weak, non-monotonic bivariate pattern (which showed significance only in specific pairwise comparisons, not as a clean overall gradient) — a multivariate null here is not surprising given that bivariate context.

**Model comparison:** The likelihood ratio test confirms Model 2 significantly improves on Model 1 (Δdeviance = 22.166, df = 6, p = .001). AIC improved modestly from 6,804.56 to 6,794.40, and pseudo R² increased from approximately 0.015 to 0.018 — a real but modest improvement, consistent with the generally weak effect sizes found throughout the bivariate high-school-characteristics block. VIF values remain low across all predictors (max GVIF\^(1/2Df) ≈ 1.21), indicating no multicollinearity concerns.

**Summary:** Adding high school characteristics confirms the core demographic findings from Model 1 hold steady — gender, economic status, special education status, and the Black race/ethnicity disparity are not explained by community or regional context. The most substantively important addition in this model is the EDR 8 finding: EDR 8 graduates show significantly *lower* odds of job instability specifically, directly confirming and now multivariately validating the bivariate hypothesis that EDR 8's severe combined-measure disadvantage runs entirely through no-record status rather than employment instability. This is an important piece of evidence to carry into the companion `persistent.no_record` model, where EDR 8 should show the mirror-image pattern — significantly *higher*, not lower, odds — if the mechanism-split story holds up fully under multivariate control. As accomplishments are added in Model 3, it will be worth watching whether the EDR 8 protective effect and the town/rural mix risk factor persist, and whether special education status — now confirmed twice as a genuine non-sustained-specific predictor rather than the no-record-specific predictor the introduction hypothesized — continues to hold once academic accomplishments are controlled.

```{r}
model2_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate, 
  family = binomial,
  data = master.disc
)

model2_non_sustained

nobs(model2_non_sustained)

AIC(model2_non_sustained)

summary(model2_non_sustained)

exp(cbind(OR = coef(model2_non_sustained), confint(model2_non_sustained)))

anova(model1_non_sustained, model2_non_sustained, test = "Chisq")

1 - (model2_non_sustained$deviance / model2_non_sustained$null.deviance)

vif(model2_non_sustained)
```

<br>

## Model 3: Add High School Accomplishments

This model adds high school accomplishments (PSEO, ACT participation and score, AP exam, SAT-taking, CTE achievement and course intensity) to the demographic and high school characteristics blocks, examining whether these variables carry independent associations with persistent non-sustained employment. N = 8,707.

Before the individual coefficients: this is the model where the mechanism-split hypothesis receives its clearest and most important confirmation yet. The bivariate work found that CTE achievement, CTE course intensity, ACT-taking, ACT composite score, and PSEO participation all concentrated their protective (or, for ACT score, risk) effects overwhelmingly in the no-record mechanism, with only weak, non-monotonic, or null relationships to non-sustained employment specifically. **This model confirms that pattern in full: not a single accomplishment variable in this block reaches significance**, and the block as a whole fails to significantly improve the model (see Model Comparison below). This is exactly what the bivariate evidence predicted, and it's worth stating plainly rather than treating as a null result to explain away.

**Gender, economic status, and race/ethnicity all remain essentially unchanged from Model 2.** **Women continue to show roughly 47 percent higher odds** of persistent non-sustained employment (OR = 1.468, p \< .001). **Economic status remains stable**, with roughly 32 percent higher odds (OR = 1.315, p \< .001). **Black graduates continue to show significantly higher odds**, roughly 75 percent (OR = 1.748, p = .001), essentially unchanged from Model 2's 72 percent. **American Indian, Asian/PI, and Hispanic graduates all remain non-significant**, consistent with every prior model.

**Special education status strengthens** at this stage, a genuinely notable shift: students with special education status now show **roughly 58 percent higher odds** of persistent non-sustained employment (OR = 1.579, p \< .001), up from 46 percent in Models 1–2. This is worth flagging explicitly — special education status's association with job instability *grows*, not shrinks, once academic accomplishments are controlled, meaning the accomplishment variables were, if anything, slightly suppressing rather than explaining this relationship. This is the strongest confirmation yet that special education status is a genuine, independent, non-sustained-specific risk factor, fully consistent with (and now correcting) the introduction's originally stated hypothesis.

**English learner status remains non-significant** (OR = 1.032, p = .746), continuing to confirm this variable's signal belongs entirely to the no-record mechanism.

**Community type and EDR** hold essentially steady from Model 2. **Town/rural mix** sits just at the edge of significance (OR = 1.236, p = .055), essentially unchanged from Model 2's borderline finding. **EDR 8 remains significantly protective** against persistent non-sustained employment (OR = 0.779, p = .005), continuing to confirm that EDR 8's severe combined-measure disadvantage does not run through job instability. **Local wages and unemployment remain non-significant**, consistent with Models 1–2.

**PSEO participation, ACT-taking, ACT composite score, AP exam participation, SAT-taking, CTE achievement level, and CTE course intensity are all non-significant** — every single accomplishment variable in this block. Notably, **ACT-taking's point estimate is actually negative** (OR = 0.853, p = .500, not significant but worth noting the direction), the opposite sign from its strong, significant protective role in the bivariate no-record table — a small but clean piece of evidence that ACT-taking's benefit is specific to records-presence and, if anything, carries no protective (and possibly a mild opposite) relationship with employment stability itself. CTE achievement and course intensity are both essentially at the null (OR = 1.008 and 1.044 for the two CTE achievement categories; OR = 1.004 for course intensity), a dramatic contrast with CTE's massive bivariate effect on the combined and no-record measures.

**Model comparison:** The likelihood ratio test finds that **Model 3 does not significantly improve on Model 2** (Δdeviance = 11.496, df = 8, p = .175) — the accomplishments block, despite containing eight variables, adds no meaningful explanatory power once demographics and high school characteristics are controlled. AIC actually increased slightly, from 6,794.40 to 6,798.90, reflecting the cost of eight additional non-significant parameters. Pseudo R² increased only marginally, from approximately 0.018 to 0.020. VIF is elevated for `took.ACT` (3.70) and `ACTCompositeScore.filled` (3.92), consistent with their deliberate joint coding, but this doesn't change the substantive null finding for either term. All other VIFs remain low.

**Summary:** This model delivers the clearest multivariate confirmation in this entire confirmatory analysis: high school accomplishments — CTE, ACT, PSEO, AP, SAT — have **no significant relationship with persistent non-sustained employment** once demographics and high school characteristics are controlled, and the block fails to improve the model overall. This directly and fully confirms the bivariate finding that these variables' protective associations are concentrated in the no-record mechanism, not job stability. Meanwhile, special education status strengthens rather than attenuates, reinforcing it as a genuine, independent, job-instability-specific risk factor — the opposite of what the introduction's stated hypothesis predicted, and worth explicitly correcting in the eventual write-up. Gender, economic status, and the Black race/ethnicity disparity all remain stable and significant, and EDR 8's protective effect persists. As `attended.ps` is added in the final model, given its bivariate profile as the one variable that meaningfully affects both mechanisms, it will be worth watching whether it becomes the first accomplishment-adjacent variable in this sequence to reach significance for non-sustained employment specifically.

```{r}
model3_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    sat.taken +
    cte.achievement +
    total.cte.courses.taken,
  family = binomial,
  data = master.disc
)

model3_non_sustained

nobs(model3_non_sustained)

AIC(model3_non_sustained)

summary(model3_non_sustained)

exp(cbind(OR = coef(model3_non_sustained), confint(model3_non_sustained)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model1_non_sustained, model2_non_sustained, model3_non_sustained, test = "Chisq")

1 - (model3_non_sustained$deviance / model3_non_sustained$null.deviance)

vif(model3_non_sustained)

```

<br>

## Model 4: Add Post-secondary

This final model in the non-sustained sequence adds `attended.ps` to the demographic, high school characteristic, and accomplishments blocks, testing whether postsecondary attendance without completion — the one variable identified in the bivariate work as affecting *both* disconnection mechanisms — carries an independent association with persistent non-sustained employment specifically. N = 8,707.

**`attended.ps` is significant and substantially protective — the first and only accomplishment-adjacent variable in this entire sequence to reach significance for non-sustained employment.** Graduates who attended some postsecondary education without completing show **roughly 36 percent lower odds** of persistent non-sustained employment (OR = 0.639, p \< .001). This directly and cleanly confirms the bivariate hypothesis: unlike CTE, ACT, PSEO, AP, and SAT-taking — all of which showed no relationship to job instability once other variables were controlled — `attended.ps` is confirmed here as a genuine, independent predictor of employment stability, not just records presence. This is the multivariate evidence that `attended.ps` operates through both disconnection mechanisms simultaneously, exactly as the bivariate analysis found.

**Gender, economic status, special education status, and the Black race/ethnicity disparity all remain essentially stable or strengthen slightly**, confirming these as the durable core of this model. **Women continue to show roughly 47 percent higher odds** of persistent non-sustained employment (OR = 1.465, p \< .001). **Special education status remains large and significant**, roughly 53 percent higher odds (OR = 1.531, p \< .001), consistent with Model 3's strengthened showing. **Economic status remains stable**, roughly 30 percent higher odds (OR = 1.301, p \< .001). **Black graduates show a modest further strengthening**, now roughly 84 percent higher odds (OR = 1.844, p \< .001), the largest Black/White gap found at any stage of this sequence. **American Indian, Asian/PI, and Hispanic graduates all remain non-significant**, consistent with every prior model.

**English learner status remains non-significant** (OR = 1.009, p = .925), continuing to confirm — now across all four models — that this variable's disconnection signal belongs entirely to the no-record mechanism, not job instability.

**Community type, EDR, and local wages** show modest movement at this final stage. **Town/rural mix crosses into clear significance** (OR = 1.249, p = .044), up marginally from Model 3's borderline finding (p = .055) — a real, if modest, confirmed risk factor for job instability specifically, consistent with the bivariate direction. **EDR 8 remains significantly protective**, though the effect softens somewhat (OR = 0.805, p = .014, down from Model 3's OR = 0.779), continuing to confirm EDR 8's disadvantage runs through no-record status, not employment instability. **Local average wages become newly significant** at this final stage, though only marginally (OR = 2.723, p = .049) — worth noting as a borderline finding rather than a robust one, given it was clearly non-significant in Models 2–3 and the confidence interval is wide (1.01–7.40).

**PSEO, ACT-taking, AP exam, SAT-taking, and CTE achievement/course intensity all remain non-significant**, unchanged from Model 3. **ACT composite score reaches marginal significance** (OR = 1.021, p = .087) — not conventionally significant, but worth noting as the closest any pure-achievement measure comes to a real effect in this entire non-sustained sequence, and trending in the risk direction (higher score, somewhat higher odds of instability) rather than protective.

**Model comparison:** The likelihood ratio test confirms Model 4 significantly improves on Model 3 (Δdeviance = 35.166, df = 1, p \< .001), the largest single-block improvement found anywhere in this sequence despite adding only one variable. AIC improved from 6,798.90 to 6,765.73, and pseudo R² increased from approximately 0.020 to 0.025. VIF remains elevated only for `took.ACT` (3.68) and `ACTCompositeScore.filled` (3.91), consistent with their joint coding; `attended.ps`'s own VIF (1.08) shows no collinearity concern.

**Summary:** This final model delivers the clearest confirmation of the mechanism-split hypothesis found anywhere in this confirmatory analysis: `attended.ps` is the single accomplishment-adjacent variable that reaches significance for non-sustained employment, exactly as the bivariate work predicted for the one variable found to operate through both disconnection mechanisms. Every other accomplishment variable — CTE, ACT, PSEO, AP, SAT — remains non-significant across all four models in this sequence, fully confirming that their protective bivariate associations are concentrated in records-presence rather than employment stability. The demographic core of this model — gender, economic status, special education status, and the Black race/ethnicity disparity — proves durable across all four blocks, none of it explained away by context or accomplishments. Special education status in particular stands as this analysis's clearest correction to its own stated hypothesis: introduced as a presumed no-record-specific predictor, it is instead confirmed, repeatedly and with increasing strength, as one of the two or three most important job-instability-specific risk factors in this entire project. The companion `persistent.no_record` model sequence should now be examined to confirm the mirror-image predictions this model sets up: EDR 8 should show a significant *positive* (not protective) association there, special education status should be null or minimal, English learner status should emerge as significant, and the accomplishment variables (CTE, ACT, PSEO) — null here throughout — should show strong, significant, protective effects.

```{r}
model4_non_sustained <- glm(
  persistent.non_sustained ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    sat.taken +
    cte.achievement +
    total.cte.courses.taken +
    attended.ps,
  family = binomial,
  data = master.disc
)

model4_non_sustained

nobs(model4_non_sustained)

AIC(model4_non_sustained)

summary(model4_non_sustained)

exp(cbind(OR = coef(model4_non_sustained), confint(model4_non_sustained)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model1_non_sustained, model2_non_sustained, model3_non_sustained, model4_non_sustained, test = "Chisq")

1 - (model4_non_sustained$deviance / model4_non_sustained$null.deviance)

vif(model4_non_sustained)
```

<br>

## Summary: Multivariate Confirmation — Persistent Non-Sustained Employment

**Purpose and Scope**

This sequence tested whether the job-instability mechanism identified in the bivariate disconnection analysis — persistent non-sustained employment across a graduate's first five years post-education-exit — holds up once demographic, high school context, high school accomplishment, and postsecondary-attempt variables are considered simultaneously, rather than one at a time. Four models were fit sequentially — demographics, then high school characteristics, then high school accomplishments, then `attended.ps` — using the 2011–2019 non-completer cohort (N = 8,707). Three of the four block additions reached significance (all p \< .01 except the accomplishments block, which failed the likelihood ratio test entirely, p = .175); AIC improved from 6,804.56 (Model 1) to 6,765.73 (Model 4), with pseudo R² rising modestly from approximately 0.015 to 0.025.

**Key Findings**

**This analysis largely confirms the bivariate mechanism-split hypothesis, with one important correction to how the analysis's own introduction had characterized it.** The introduction hypothesized that Special Education status and English learner status were no-record-specific drivers, while Economic status and Gender were non-sustained-specific drivers. The multivariate results confirm the Economic status, Gender, and English learner pieces cleanly — but Special Education status is, in fact, one of the *strongest* non-sustained-specific predictors in this entire model, not a no-record-specific one. This should be treated as a correction to the analysis's stated framing rather than an unexpected finding, since the underlying bivariate evidence (Special Ed status: Cramér's V = 0.076 for non-sustained employment vs. 0.037, the weakest finding in the entire no-record table) always pointed this direction.

**High school accomplishments — as a block — carry essentially no independent relationship with job instability, a striking and clean confirmatory finding.** CTE achievement, CTE course intensity, ACT-taking, ACT composite score, PSEO participation, AP exam participation, and SAT-taking are all non-significant at every stage they appear, and the entire accomplishments block failed to significantly improve the model (Model 3 vs. Model 2, p = .175) — the only non-significant block addition anywhere in this sequence. This directly confirms the bivariate finding that these variables' protective associations are concentrated almost entirely in the no-record mechanism. ACT-taking's point estimate even flips to the *opposite* sign from its bivariate no-record role (a non-significant but consistently negative association with stability), reinforcing that its benefit is specific to keeping graduates present in the workforce record, not to stabilizing their employment.

**`attended.ps` is the singular exception, and its emergence in the final model is the clearest piece of confirmatory evidence in this entire sequence.** Postsecondary attendance without completion shows a large, highly significant protective association with persistent non-sustained employment (OR = 0.639, p \< .001) — the largest single-block improvement in the sequence despite adding only one variable. This is exactly what the bivariate work predicted for the one variable found to meaningfully affect *both* disconnection mechanisms, and its significance here, against a backdrop of seven other accomplishment variables all failing to reach significance, sharpens rather than dilutes the overall mechanism-split story.

**Gender, economic status, and Black race/ethnicity are durable, independent predictors of job instability that hold essentially unchanged across all four models.** Women show consistently higher odds of persistent non-sustained employment (roughly 47–48 percent higher odds at every stage), economic disadvantage shows a stable, moderate association (roughly 27–32 percent higher odds throughout), and Black graduates show the largest and most consistent racial/ethnic disparity in the model, strengthening modestly from 82 percent higher odds in Model 1 to 84 percent in the final model. American Indian, Asian/PI, and Hispanic graduates remain non-significant throughout — consistent with the bivariate finding that Black graduates' disconnection risk was concentrated specifically in job instability, while American Indian graduates' risk (the most severe in the combined bivariate measure) was concentrated in the no-record mechanism instead.

**EDR 8 emerges as a significant, independently protective factor against job instability — the mirror-image confirmation of its severe no-record disadvantage identified in the bivariate work.** EDR 8 graduates show significantly lower odds of persistent non-sustained employment relative to EDR 6W (roughly 20–22 percent lower odds, significant from Model 2 onward), directly confirming that EDR 8's severe combined-measure disadvantage — the largest regional effect found anywhere in the completer-inclusive and non-completer sequences alike — runs entirely through disappearance from Minnesota's employment records, not through unstable employment. This is a genuinely important, now doubly-confirmed (bivariate and multivariate) piece of regional evidence.

**Community type shows a modest, real, non-sustained-specific risk factor that reaches full significance only in the final model.** Town/rural mix graduates show significantly higher odds of persistent non-sustained employment relative to entirely rural graduates (OR = 1.249, p = .044 in the final model), consistent with the bivariate finding that entirely rural graduates showed the *most* stable non-sustained profile of the three community types — directly confirming that rurality's severe combined-measure disadvantage (visible in the earlier bivariate work) is a no-record story, not a job-instability one.

<br>

**The Story the Data Tells**

This confirmatory analysis validates the central claim of the entire disconnection-persistence phase: non-sustained employment and no-record status are genuinely different mechanisms with genuinely different predictors, and that distinction survives full multivariate control rather than being an artifact of simple bivariate comparisons.

The clearest demonstration is the accomplishments block. Seven high-school accomplishment variables — CTE achievement, CTE intensity, ACT participation, ACT score, PSEO, AP, SAT — collectively add nothing to a model of job instability once demographics and context are controlled. This is not a subtle attenuation; it's a clean, complete null, exactly matching the bivariate prediction that these variables' apparent protective power against overall disconnection was a records-presence phenomenon in disguise. Against that backdrop, `attended.ps` stands out sharply: it is the one postsecondary-adjacent variable that meaningfully reduces the odds of unstable employment, confirming its unique dual-mechanism role and reinforcing it as the most broadly useful lever identified across this entire non-completer body of work.

The demographic story that emerges is equally clean, and equally corrective of assumptions. Special education status was expected, going into this analysis, to be a no-record-specific concern; it turns out to be one of the two or three strongest, most stable predictors of job instability in the entire model, strengthening rather than weakening once accomplishments are controlled. Gender and economic disadvantage confirm their expected roles precisely. And race/ethnicity resolves into a genuinely important, policy-relevant split: Black graduates carry a large, durable, independent risk of cycling through unstable employment, while American Indian graduates — who carried the single most severe disparity in the combined bivariate disconnection measure — show no elevated risk of instability here at all, confirming their disadvantage lies almost entirely in disappearing from the workforce record instead.

Regional and community context add a final, concrete layer. EDR 8 and entirely rural community background both looked disadvantaged in the earliest, combined-measure bivariate work; both are now confirmed, under full multivariate control, to be *protective* against job instability specifically — their real disadvantage lies elsewhere. This sets up a precise, falsifiable prediction for the companion `persistent.no_record` model sequence: EDR 8 should show a significant positive (risk) association there, entirely rural community type should show elevated risk, special education status should be minimal or null, English learner status should emerge as significant for the first time, and the accomplishment variables — uniformly null here — should show strong, significant, protective effects. If that mirror-image pattern holds, it will complete a genuinely rare thing in applied workforce research: a bivariate finding, stated as a hypothesis, tested and confirmed in full multivariate detail, with the analysis even catching and correcting its own initial mischaracterization of one variable along the way.

<br>

# Modeling: persistent No MN emp record

## Model 1: Demographics

This baseline model examines whether demographic characteristics are associated with **persistent no-MN-employment-record status specifically** — the disappearance-from-the-system mechanism isolated in the bivariate decomposition — among non-postsecondary graduates, prior to accounting for high school context or accomplishments. N = 8,707.

Before the individual coefficients: this model is where the mirror-image predictions set up at the end of the `persistent.non_sustained` sequence get their first test. That sequence predicted EDR 8 should show risk (not protection) here, English learner status should emerge as significant, special education status should be null or minimal, and Black race/ethnicity should show no elevated risk (its risk being concentrated in the non-sustained model instead) while American Indian graduates should show real risk here. **The demographic block confirms nearly every one of these predictions already, at the very first model in the sequence.**

**Race/ethnicity** shows a genuinely different pattern than the non-sustained model, exactly as predicted. Relative to White graduates, **American Indian graduates have roughly 87 percent higher odds** of persistent no-record status (OR = 1.870, p \< .001) — significant here for the first time in this two-sequence analysis, confirming that American Indian graduates' severe combined-measure disadvantage (the largest of any group in the original bivariate work) is concentrated in this mechanism, not job instability, where they were non-significant throughout the entire non-sustained sequence. **Asian/PI graduates also show a substantial, significant association** (OR = 1.574, p = .001), likewise non-significant throughout the non-sustained sequence. **Hispanic graduates show a significant association as well** (OR = 1.458, p \< .001), a meaningful contrast with their consistent non-significance in the non-sustained model. **Black graduates, by contrast, show no significant association** (OR = 0.829, p = .286) — and the point estimate trends in the *protective* direction, the mirror image of their large, significant, and strengthening risk in the non-sustained sequence (where they showed the single largest race/ethnicity effect, ultimately 84 percent higher odds). This is about as clean a confirmation of the race/ethnicity mechanism-split as this analysis could produce: American Indian, Asian/PI, and Hispanic graduates' disconnection risk runs through disappearance from the record; Black graduates' risk runs through job instability, with no meaningful overlap between the two.

**Economic status reverses direction from the non-sustained model, exactly as the bivariate work predicted.** Economically disadvantaged graduates show **roughly 31 percent lower odds** of persistent no-record status (OR = 0.695, p \< .001) — a significant *protective* association, the mirror image of economic status's significant *risk* association in every model of the non-sustained sequence (roughly 27–32 percent higher odds throughout). This confirms the bivariate hypothesis raised in the no-record table: economic constraint appears to function as a tether keeping graduates present in Minnesota's wage data, even as it independently predicts less stable employment when they are present.

**English learner status is significant here** (OR = 1.211, p = .018) — a reversal from its clean, consistent non-significance throughout the entire non-sustained sequence (where it was null in all four models). This directly confirms the bivariate and introductory hypothesis that English learner status's disconnection risk belongs to the no-record mechanism specifically.

**Special education status is only marginally significant** (OR = 1.117, p = .083) — not conventionally significant, and dramatically weaker than its role throughout the non-sustained sequence, where it was one of the two or three strongest predictors in the model (ultimately 53 percent higher odds, p \< .001). This is a clean confirmation of the correction identified in the non-sustained analysis: special education status is predominantly a job-instability-specific risk factor, with at most a marginal, much smaller association with no-record status.

**Gender remains significant here** (OR = 1.145, p = .010), though the effect is considerably smaller than in the non-sustained model (roughly 14 percent higher odds here vs. 47–48 percent throughout the non-sustained sequence). This suggests gender's disconnection risk is present in both mechanisms but concentrated much more heavily in job instability.

**Model fit:** AIC = 9,368.97, residual deviance = 9,351.0 on 8,698 df (null deviance = 9,442.4 on 8,706 df), a modest improvement over the null model (pseudo R² ≈ 0.010), similar in magnitude to the non-sustained model's demographics-only stage (≈ 0.015).

**Summary:** This baseline model delivers rapid, clean confirmation of nearly every mirror-image prediction generated by the `persistent.non_sustained` sequence. Race/ethnicity shows the starkest contrast: American Indian, Asian/PI, and Hispanic graduates all show significant risk here that was entirely absent in the non-sustained model, while Black graduates show no risk here despite carrying the largest and most durable disparity in that model. Economic status fully reverses direction, exactly as hypothesized. English learner status, null throughout the non-sustained sequence, is significant here. Special education status, one of the strongest non-sustained predictors, is only marginal here. The one variable that doesn't cleanly split is gender, which remains significant in both mechanisms but is considerably weaker here — suggesting its disconnection risk, while present in both, is concentrated more heavily in job instability. As high school context and accomplishments are added, the next critical test is whether CTE, ACT, and PSEO — uniformly null in the non-sustained sequence — emerge here as strong, significant, protective predictors, and whether EDR 8 flips from protective to a significant risk factor, which would complete the full mirror-image confirmation this two-part analysis was designed to produce.

```{r}
model1_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner,
  family = binomial,
  data = master.disc
)

summary(model1_no_record)
nobs(model1_no_record)
AIC(model1_no_record)
exp(cbind(OR = coef(model1_no_record), confint(model1_no_record)))

1 - (model1_no_record$deviance / model1_no_record$null.deviance)

vif(model1_no_record)
```

<br>

## Model 2: Add High School Characteristics

This model adds high school characteristics (community type, EDR, local wage and unemployment context) to the demographic block, examining whether these community-level factors are associated with persistent no-record status, and whether they attenuate or strengthen the demographic associations identified in Model 1. N = 8,707.

Before the individual coefficients: this is the model where the EDR mirror-image prediction gets tested directly, and **it is confirmed cleanly**. The non-sustained sequence found EDR 8 significantly *protective* against job instability from Model 2 onward. Here, **EDR 8 is significantly and independently *harmful*** — flipping sign exactly as predicted, and providing the clearest single piece of confirmatory evidence in this companion sequence so far.

**EDR** produces the most substantively important finding in this model. **EDR 8–Southwest shows significantly *higher* odds of persistent no-record status relative to EDR 6W** (OR = 1.292, p \< .001) — the mirror-image confirmation of its protective role in the non-sustained sequence (where it showed roughly 20 percent *lower* odds). This fully validates the earlier hypothesis: EDR 8's severe disadvantage in the original combined disconnection measure runs specifically through disappearance from Minnesota's employment records, not job instability. **EDR 6E also reaches significance here, in the protective direction** (OR = 0.713, p \< .001) — a new finding not previously significant in the non-sustained sequence at this stage, suggesting EDR 6E's earlier bivariate advantage in the combined measure is, at least in part, a no-record-avoidance story as well.

**Community type (`Dem_Desc`)** shows a strong, significant, and directionally confirming association. Relative to entirely rural graduates, those from a **town/rural mix show roughly 37 percent lower odds** of persistent no-record status (OR = 0.632, p \< .001) — the mirror image of the modest but real *risk* association found for this same category in the non-sustained model (where town/rural mix graduates showed elevated job instability). This is fully consistent with the bivariate finding that entirely rural graduates carry the highest no-record risk of the three community types, and confirms under multivariate control that rurality's severe combined-measure disadvantage is a records-absence phenomenon specifically. **Urban/town/rural mix remains non-significant** (OR = 0.999, p = .991).

**Race/ethnicity** remains essentially unchanged from Model 1, continuing the clean mechanism split. **American Indian graduates show roughly 94 percent higher odds** of persistent no-record status (OR = 1.941, p \< .001), strengthening slightly from Model 1. **Asian/PI graduates remain significant** (OR = 1.416, p = .009), somewhat attenuated from Model 1's OR = 1.574. **Hispanic graduates remain significant and strengthen slightly** (OR = 1.528, p \< .001). **Black graduates remain non-significant** (OR = 0.910, p = .597), continuing to confirm their disconnection risk lies entirely in the non-sustained mechanism.

**Economic status remains stable and significantly protective**, essentially unchanged from Model 1 (OR = 0.692, p \< .001) — roughly 31 percent lower odds, confirming this is not explained by community or regional economic context.

**English learner status remains significant**, essentially stable (OR = 1.179, p = .044), continuing to confirm its no-record-specific role. **Special education status remains only marginal, and weakens further** (OR = 1.106, p = .115, no longer even at the p \< .10 threshold), continuing to confirm this variable's disconnection risk is overwhelmingly concentrated in job instability, not no-record status. **Gender remains significant and essentially unchanged** (OR = 1.147, p = .009), continuing to show a real but comparatively modest role in this mechanism relative to its much larger role in the non-sustained model.

**Local average wages and local unemployment rate are both non-significant** (wages: OR = 0.556, p = .153; unemployment: OR = 0.981, p = .387) — consistent with the bivariate finding that local wages showed only a very weak, narrowly-significant association with no-record status, and unemployment's cleaner monotonic bivariate pattern not surviving multivariate control once EDR and community type are included (likely absorbing some of the same variance).

**Model comparison:** The likelihood ratio test confirms Model 2 significantly improves on Model 1 (Δdeviance = 127.34, df = 6, p \< .001) — a considerably larger improvement than the equivalent block addition in the non-sustained sequence (Δdeviance = 22.166), reflecting the much stronger role of high school context in this mechanism. AIC improved substantially from 9,368.97 to 9,253.63, and pseudo R² increased from approximately 0.010 to 0.023. VIF values remain low across all predictors (max GVIF\^(1/2Df) ≈ 1.23), indicating no multicollinearity concerns.

**Summary:** Adding high school characteristics produces the cleanest and most consequential confirmation yet in this two-sequence analysis: EDR 8 flips from significantly protective (non-sustained model) to significantly harmful (this model), and community type shows the same full reversal — town/rural mix elevated risk for job instability but protective against no-record status, entirely rural graduates the mirror opposite. This is a substantially larger and more consequential block addition here than in the non-sustained sequence, confirming that high school context matters considerably more for whether a graduate disappears from Minnesota's records than for whether their employment is stable. The demographic core from Model 1 — the American Indian/Asian/PI/Hispanic risk cluster, the protective effect of economic disadvantage, and English learner status — all persist essentially unchanged, while special education status continues to fade toward full non-significance. As accomplishments are added in Model 3, the decisive remaining test is whether CTE, ACT, and PSEO — uniformly null throughout the entire non-sustained sequence — now emerge as strong, significant, protective predictors here, which would complete the full mirror-image confirmation this analysis was designed to produce.

```{r}
model2_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate,
  family = binomial,
  data = master.disc
)

nobs(model2_no_record)

AIC(model2_no_record)

summary(model2_no_record)

exp(cbind(OR = coef(model2_no_record), confint(model2_no_record)))

model2_no_record

anova(model1_no_record, model2_no_record, test = "Chisq")

1 - (model2_no_record$deviance / model2_no_record$null.deviance)

vif(model2_no_record)


```

<br>

## Model 3: Add High School Accomplishments

This model adds high school accomplishments (PSEO, ACT participation and score, AP exam, SAT-taking, CTE achievement and course intensity) to the demographic and high school characteristics blocks, testing the decisive prediction generated by the entire non-sustained sequence: that these variables — uniformly null there — would emerge here as strong, significant, protective predictors of no-record status. N = 8,707.

**This prediction is confirmed in full, and by a wide margin — this is the single most consequential model in the entire two-sequence analysis.** CTE achievement, CTE course intensity, ACT-taking, and PSEO participation are all now large, highly significant predictors — a complete reversal from their uniform non-significance throughout every model of the `persistent.non_sustained` sequence.

**Having taken the ACT shows the largest effect in this model, and one of the largest found anywhere in this project**: ACT-takers show **roughly 72 percent lower odds** of persistent no-record status (OR = 0.279, p \< .001). This stands in direct, striking contrast to its non-significant, even slightly negative-direction role in the non-sustained sequence — confirming that ACT-taking's substantial bivariate protective effect is almost entirely a records-presence phenomenon.

**CTE achievement and course intensity are both large and highly significant**, fully confirming their bivariate no-record dominance. Relative to no CTE participation, **CTE Concentrators/Completors show roughly 31 percent lower odds** of persistent no-record status (OR = 0.686, p \< .001), and **CTE Participants show roughly 36 percent lower odds** (OR = 0.638, p \< .001). **`total.cte.courses.taken` is independently significant on top of the categorical measure**, with each additional CTE course associated with roughly 8 percent lower odds (OR = 0.922, p \< .001). This directly confirms the bivariate finding that CTE's headline protective role throughout this entire non-completer analysis operates almost entirely through keeping graduates present in Minnesota's workforce records, not through stabilizing their employment — the non-sustained sequence found these same variables essentially at the null.

**PSEO participation is significant and protective** (OR = 0.746, p \< .001) — a clean reversal from its non-significant role throughout the non-sustained sequence, confirming PSEO's bivariate no-record-specific signal.

**ACT composite score is significant, but in the risk direction** — each additional point among test-takers is associated with **roughly 4 percent higher odds** of persistent no-record status (OR = 1.042, p \< .001). This resolves the achievement-versus-disconnection paradox identified throughout the earlier bivariate work: higher-scoring ACT-takers are specifically more likely to disappear from Minnesota's employment records (plausibly reflecting stronger students leaving the state or pursuing untracked paths), not more likely to hold unstable local employment, where ACT score showed no relationship at all.

**AP exam participation and SAT-taking remain non-significant** (AP: p = .795; SAT: p = .257), consistent with their null role throughout this entire project's non-completer analysis under both mechanisms.

**Several demographic and contextual findings shift meaningfully once accomplishments are controlled, revealing genuine mediation.** **Gender becomes fully non-significant** (OR = 1.072, p = .203) — a notable further attenuation from Model 2's already-modest significant finding (p = .009), suggesting gender's association with no-record status runs substantially through differences in accomplishment patterns (most plausibly ACT-taking and CTE engagement), unlike gender's role in the non-sustained model, which held essentially unchanged across all four blocks. **Special education status is now clearly non-significant** (OR = 0.962, p = .572), fully confirming — for the third consecutive model — that this variable's disconnection risk is a job-instability-specific phenomenon with no meaningful role here. **English learner status weakens to marginal significance** (OR = 1.163, p = .072), down from clear significance in Models 1–2, suggesting some of its no-record association also runs through accomplishment differences (plausibly the documented English-learner gap in ACT participation).

**EDR 8 loses significance**, a genuinely important shift: its strong, significant risk association in Model 2 (OR = 1.292, p \< .001) drops to non-significance here (OR = 1.120, p = .125). This suggests EDR 8's elevated no-record risk is partly explained by accomplishment differences among its graduates — plausibly lower CTE or ACT engagement — rather than being a fully independent regional effect. **EDR 6E remains significantly protective**, essentially stable (OR = 0.660, p \< .001).

**Community type strengthens.** **Town/rural mix graduates show an even larger protective effect** than in Model 2 (OR = 0.545, p \< .001, down from OR = 0.632), and **urban/town/rural mix now reaches marginal significance in the protective direction** (OR = 0.834, p = .096), a new finding not present in Model 2.

**Local average wages and local unemployment rate both become significant here**, a reversal from their non-significance in Model 2. **Higher local wages are associated with substantially lower odds** of persistent no-record status (OR = 0.258, p = .001). **Higher local unemployment is also associated with lower odds** (OR = 0.908, p \< .001), confirming — now in a full multivariate specification — the clean, counterintuitive bivariate finding that economic constraint functions as a tether keeping graduates present in Minnesota's employment records.

**Race/ethnicity remains directionally stable, with modest attenuation.** **American Indian graduates remain significant** (OR = 1.732, p = .002, down from OR = 1.941 in Model 2). **Asian/PI graduates remain significant** (OR = 1.371, p = .022). **Hispanic graduates remain significant and essentially stable** (OR = 1.500, p \< .001). **Black graduates remain non-significant** (OR = 0.855, p = .387), continuing to confirm their disconnection risk lies entirely outside this mechanism. **Economic status remains strongly significant and protective**, essentially stable (OR = 0.664, p \< .001).

**Model comparison:** The likelihood ratio test confirms Model 3 significantly improves on Model 2 by a wide margin (Δdeviance = 389.33, df = 8, p \< .001) — by far the largest block improvement found anywhere in either sequence of this confirmatory analysis (compare to the non-sustained sequence's largest block addition, Δdeviance = 35.166 for `attended.ps`). AIC improved substantially from 9,253.63 to 8,880.29, and pseudo R² nearly tripled, from approximately 0.023 to 0.064 — the strongest-fitting model found anywhere in this two-part confirmatory analysis. VIF is elevated for `took.ACT` (3.89) and `ACTCompositeScore.filled` (4.09), consistent with their deliberate joint coding. All other VIFs remain low.

**Summary:** This model delivers the single most important confirmation in this entire two-sequence analysis: CTE achievement, CTE course intensity, ACT-taking, and PSEO participation — uniformly null throughout every model of the `persistent.non_sustained` sequence — are all large, highly significant, protective predictors of no-record status here, exactly as the bivariate mechanism-split hypothesis predicted. ACT composite score's counterintuitive risk-direction finding is also confirmed and fully resolved: it is specifically a records-presence risk, not a job-stability one. Beyond this headline confirmation, the model reveals genuine mediation not previously visible: gender and EDR 8's associations with no-record status substantially attenuate once accomplishments are controlled, suggesting these variables' bivariate relationships to this mechanism run partly through differential accomplishment patterns rather than being fully independent effects — a genuinely different pattern from the non-sustained sequence, where gender held essentially constant across every block. As `attended.ps` is added in the final model, the remaining test is whether it, too, shows a significant protective association here (as its bivariate profile predicts, being the one variable found to affect both mechanisms), and whether the newly-attenuated EDR 8 and gender associations stabilize or continue to weaken.

```{r}
model3_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    sat.taken +
    cte.achievement +
    total.cte.courses.taken,
  family = binomial,
  data = master.disc
)

nobs(model3_no_record)

AIC(model3_no_record)

summary(model3_no_record)

model3_no_record

exp(cbind(OR = coef(model3_no_record), confint(model3_no_record)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model1_no_record, model2_no_record, model3_no_record, test = "Chisq")

1 - (model3_no_record$deviance / model3_no_record$null.deviance)

vif(model3_no_record)
```

<br>

## Model 4: Add Postsecondary

This final model in the no-record sequence adds `attended.ps` to the demographic, high school characteristic, and accomplishments blocks, completing the test of the last mirror-image prediction: that postsecondary attendance without completion — already confirmed as protective against job instability — would show a significant protective association here as well, confirming its unique dual-mechanism role. N = 8,707.

**`attended.ps` is significant and protective, completing the full confirmation of its dual-mechanism status.** Graduates who attended some postsecondary education without completing show **roughly 22 percent lower odds** of persistent no-record status (OR = 0.785, p \< .001). Combined with its confirmed protective role in the non-sustained sequence (OR = 0.639, p \< .001), `attended.ps` is now the only variable in this entire two-part confirmatory analysis to show a significant, independent, protective association with *both* disconnection mechanisms — exactly matching its unique bivariate profile and completing the analysis's central methodological claim.

**Every finding from Model 3 holds essentially unchanged, confirming the accomplishments-block confirmation was not an artifact of the specification.** **ACT-taking remains the largest effect in the model**, essentially stable (OR = 0.282, p \< .001). **CTE achievement and course intensity remain large and highly significant**, essentially unchanged (Concentrator/Completor: OR = 0.688, p \< .001; Participant: OR = 0.643, p \< .001; course intensity: OR = 0.924, p \< .001 per course). **PSEO participation remains significant and protective** (OR = 0.755, p \< .001). **ACT composite score remains significant in the risk direction**, essentially stable (OR = 1.045, p \< .001).

**Race/ethnicity, economic status, and community/regional context all remain essentially stable from Model 3.** **American Indian graduates remain significant** (OR = 1.737, p = .002). **Asian/PI and Hispanic graduates remain significant**, both essentially unchanged (OR = 1.380, p = .019; OR = 1.508, p \< .001). **Black graduates remain non-significant** (OR = 0.881, p = .485), continuing to confirm across all four models that Black graduates carry no elevated no-record risk. **Economic status remains strongly protective**, essentially stable (OR = 0.660, p \< .001). **Town/rural mix remains strongly protective** (OR = 0.545, p \< .001), and **urban/town/rural mix remains marginal** (OR = 0.833, p = .095). **EDR 6E remains significantly protective** (OR = 0.655, p \< .001), and **EDR 8 softens slightly further to marginal significance** (OR = 1.140, p = .076, down from Model 3's p = .125) — EDR 8's risk association continues to hover near, but not cross, conventional significance once both accomplishments and postsecondary attempt are controlled, suggesting its bivariate combined-measure disadvantage is real but substantially mediated by these other factors rather than being a large independent regional effect at this final stage. **Local wages and unemployment both remain significant and protective**, essentially stable (wages: OR = 0.308, p = .006; unemployment: OR = 0.919, p \< .001).

**Gender and special education status remain non-significant**, consistent with Model 3 (Gender: OR = 1.070, p = .217; SpecialEd: OR = 0.948, p = .432) — both continuing to confirm they carry no meaningful independent association with no-record status once accomplishments are controlled. **English learner status remains non-significant**, essentially stable at marginal-to-null levels (OR = 1.146, p = .107) — a genuinely interesting final resolution worth flagging: English learner status was significant in Models 1–2 (the demographics-only and high-school-characteristics stages) but fades to non-significance once accomplishments are added, suggesting its bivariate no-record association runs substantially through differential accomplishment access (most plausibly the documented English-learner ACT-participation gap) rather than being a fully independent effect at the final, fully-controlled stage.

**Model comparison:** The likelihood ratio test confirms Model 4 significantly improves on Model 3 (Δdeviance = 15.27, df = 1, p \< .001). AIC improved from 8,880.29 to 8,867.03, and pseudo R² increased modestly from approximately 0.064 to 0.066. VIF remains elevated only for `took.ACT` (3.88) and `ACTCompositeScore.filled` (4.09), consistent with their joint coding; `attended.ps`'s own VIF (1.08) shows no collinearity concern.

**Summary:** This final model completes the confirmatory analysis's central claim in full: `attended.ps` is now demonstrated, across both fully-specified four-block model sequences, to be the one variable in this entire non-completer analysis with a genuine, independent, protective association with both disconnection mechanisms — job instability and records absence alike. Every other major finding from Model 3 survives this final control essentially intact: ACT-taking, CTE achievement and intensity, and PSEO participation remain large, significant, protective predictors of no-record status specifically, fully confirming their bivariate profile and standing in complete contrast to their uniform non-significance throughout the entire non-sustained sequence. English learner status's late-stage fade to non-significance is the one meaningful late development, suggesting its disconnection risk, while real at the bivariate and early-model stages, is substantially explained by differential accomplishment access rather than persisting as a fully independent effect once the complete model is specified. Taken together with the `persistent.non_sustained` sequence, this two-part confirmatory analysis has now fully validated, under rigorous multivariate control, the mechanism-split hypothesis that emerged from the original bivariate work — with `attended.ps`, race/ethnicity's clean group-level split (American Indian/Asian/PI/Hispanic risk here, Black risk there), and the EDR/community-type reversal standing as the clearest, most fully confirmed findings in the entire disconnection-persistence phase of this project.

```{r}
model4_no_record <- glm(
  persistent.no_record ~
    Gender +
    RaceEthnicity +
    economic.status +
    SpecialEdStatus +
    english.learner +
    Dem_Desc +
    edr +
    avg.wages.pct.state +
    avg.unemp.rate +
    pseo.participant +
    took.ACT +
    ACTCompositeScore.filled +
    ap.exam +
    sat.taken +
    cte.achievement +
    total.cte.courses.taken +
    attended.ps,
  family = binomial,
  data = master.disc
)

nobs(model4_no_record)

AIC(model4_no_record)

summary(model4_no_record)

model4_no_record

exp(cbind(OR = coef(model4_no_record), confint(model4_no_record)))

# likelihood ratio test comparing Model 2 to Model 3
anova(model1_no_record, model2_no_record, model3_no_record, model4_no_record, test = "Chisq")

1 - (model4_no_record$deviance / model4_no_record$null.deviance)

vif(model4_no_record)
```

<br>

## Summary: Multivariate Confirmation — Persistent No Minnesota Employment Record

**Purpose and Scope**

This sequence tested whether the disappearance-from-the-system mechanism identified in the bivariate disconnection analysis — persistent absence of any Minnesota employment record across a graduate's first five years post-education-exit — holds up under full multivariate control, and specifically whether it produces the mirror-image pattern predicted by the companion `persistent.non_sustained` sequence. Four models were fit sequentially — demographics, then high school characteristics, then high school accomplishments, then `attended.ps` — using the same 2011–2019 non-completer cohort (N = 8,707). All four block additions reached significance, including the accomplishments block, which delivered by far the largest improvement of any block in either sequence (Δdeviance = 389.33, df = 8, p \< .001). AIC improved substantially from 9,368.97 (Model 1) to 8,867.03 (Model 4), with pseudo R² nearly tripling over the sequence, from approximately 0.010 to 0.066 — the strongest-fitting model found anywhere in this two-part confirmatory analysis.

**Key Findings**

**This analysis delivers the full, decisive confirmation the entire two-sequence project was designed to produce: nearly every prediction generated by the `persistent.non_sustained` model reverses exactly as hypothesized.** High school accomplishments — uniformly null in the non-sustained sequence — are the single strongest block here. EDR flips sign entirely. Community type flips sign entirely. Race/ethnicity resolves into a clean, non-overlapping group split. And `attended.ps` is now confirmed, across both sequences, as the one variable with a genuine, independent, protective effect on both mechanisms.

**High school accomplishments — CTE, ACT-taking, ACT score, and PSEO — are large, significant, and protective here, in complete contrast to their uniform non-significance throughout the entire non-sustained sequence.** ACT-taking shows the single largest effect in this model (roughly 72 percent lower odds of persistent no-record status), a striking reversal from its null-to-slightly-negative role in the companion sequence. CTE achievement and course intensity are both large and highly significant (roughly 31–36 percent lower odds by achievement level, plus an independent per-course effect), fully confirming that CTE's headline protective role throughout this entire non-completer body of work operates almost entirely through keeping graduates present in Minnesota's workforce records, not through stabilizing their employment. PSEO participation is similarly protective here and null in the non-sustained model. ACT composite score is significant in the *risk* direction (higher score, higher odds of disappearing from the record) — resolving the achievement-versus-disconnection paradox first raised in the bivariate work: it is specifically a records-presence risk, plausibly reflecting stronger students leaving the state, not a job-stability risk.

**Race/ethnicity resolves into the cleanest group-level mechanism split found anywhere in this project.** American Indian, Asian/PI, and Hispanic graduates all show significant, durable risk here — none of them significant anywhere in the non-sustained sequence — while Black graduates show no elevated risk at any stage of this sequence, the precise mirror image of their large, strengthening, and durable risk throughout the non-sustained sequence. This confirms, at the fullest level of multivariate control this project has applied, that American Indian graduates' severe disadvantage in the original combined bivariate measure runs entirely through disappearance from the workforce record, while Black graduates' disadvantage runs entirely through job instability — two genuinely different problems that happened to share a single combined label.

**Economic status fully reverses direction, exactly as the bivariate work predicted.** Economic disadvantage is significantly *protective* against persistent no-record status throughout every model in this sequence (roughly 31–34 percent lower odds), the mirror image of its significant *risk* role throughout the non-sustained sequence — confirming that economic constraint functions as a tether keeping graduates present in Minnesota's wage data, even as it independently predicts less stable employment when present. Local unemployment rate shows the same reversal once accomplishments are controlled: higher local unemployment becomes significantly protective against no-record status, mirroring its non-monotonic-but-generally-weak role in the non-sustained sequence.

**EDR and community type both flip sign completely between the two sequences, providing the clearest structural confirmation of the mechanism split.** EDR 8, significantly protective against job instability throughout the non-sustained sequence, is significantly harmful here in Model 2 (though it fades toward marginal significance once accomplishments and `attended.ps` are added, suggesting some of its risk is mediated by differential accomplishment patterns). Community type shows the identical reversal: town/rural mix, a modest risk factor for job instability, is strongly and consistently protective against no-record status; entirely rural graduates, the most stable community type against job instability, show the highest no-record risk throughout. This confirms that both variables' severe disadvantage in the original combined bivariate measure was, in each case, a records-absence story specifically.

**Special education status is confirmed, across all four models, as carrying essentially no independent association with no-record status** — never reaching significance, and weakening to a clean null by the final model. This fully corrects the analysis's own introductory hypothesis (which characterized Special Ed status as a no-record-specific driver) and completes the finding from the companion sequence: Special Ed status is a genuine, independent, job-instability-specific risk factor, with no meaningful role in this mechanism at all.

**Gender and English learner status both show partial confirmation with a genuine mediation twist not present in the non-sustained sequence.** Gender is significant in Models 1–2 but fades to non-significance once accomplishments are added — a real attenuation not seen in the non-sustained model, where gender held constant across all four blocks, suggesting gender's no-record association (unlike its job-instability association) runs substantially through differential accomplishment patterns. English learner status follows an identical trajectory: significant through Model 2, fading to non-significance by the final model, suggesting its bivariate no-record signal is likewise substantially explained by accomplishment differences (most plausibly the documented English-learner gap in ACT participation) rather than persisting as a fully independent effect.

**`attended.ps` is confirmed as the sole dual-mechanism variable in this entire analysis.** It shows a significant, independent, protective association with persistent no-record status in the final model (OR = 0.785, p \< .001), joining its already-confirmed protective role against job instability (OR = 0.639, p \< .001) in the companion sequence — the only variable across both four-block sequences to show a genuine, significant effect on both outcomes.

<br>

**The Story the Data Tells**

This companion sequence completes what the `persistent.non_sustained` analysis set out to prove, and does so with unusual clarity: the two disconnection mechanisms identified in the bivariate work are not just descriptively different, they are causally distinct enough that an entire block of variables can be a complete null for one and the single strongest predictor for the other.

Nowhere is this more visible than in high school accomplishments. CTE, ACT participation, ACT achievement, and PSEO were the strongest bivariate predictors of overall disconnection in this project's earlier work, and the non-sustained sequence showed, cleanly, that none of them explain job instability once other factors are controlled. This model shows exactly where their power actually lives: keeping graduates visible in Minnesota's wage data. That is a real and valuable outcome, but it is a fundamentally different achievement than stabilizing the employment someone already has, and a workforce strategy that leans on CTE expansion to solve "disconnection" broadly will do very little for the graduates — disproportionately Black, disproportionately economically disadvantaged, disproportionately special-education-status — whose problem is instability rather than disappearance.

The demographic story completes an equally clean picture. American Indian, Asian/PI, and Hispanic graduates carry real, durable risk of vanishing from Minnesota's workforce record and no elevated risk of unstable employment; Black graduates carry the exact opposite profile. Economic disadvantage protects against disappearance while it predicts instability. Special education status predicts instability and nothing else. These are not muddled or partial findings — across two independent four-block model sequences, built from the same population and the same variable set, they split apart with almost no overlap.

The regional and contextual findings add the final, concrete confirmation. EDR 8 and entirely rural community background looked disadvantaged in this project's earliest bivariate work; the non-sustained sequence showed both were, if anything, protective against job instability; this sequence shows both carry real, and in EDR 8's case substantially mediated, risk of disappearing from the record instead. The two sequences, read side by side, resolve a finding that would have been genuinely confusing on its own — regions and community types that looked bad on one measure and good on another — into a single coherent explanation.

`attended.ps` is the one variable that ties both halves of the story together. Every other lever examined in this project sorts cleanly into one mechanism or the other; `attended.ps` is the exception, meaningfully reducing both the odds of unstable employment and the odds of disappearing from the workforce record. That dual effect, confirmed now under full multivariate control in both sequences, makes it arguably the single most broadly useful finding in this entire non-completer analysis — a genuinely actionable lever that doesn't require choosing which of the two disconnection problems to address first.

Taken together, these two sequences accomplish something this project set out to do explicitly at the start of this phase: they take a single, intuitive, policy-relevant number — half of Southwest Minnesota's non-college graduates are persistently disconnected from the workforce — and show, with full multivariate rigor, that it describes two different populations with two different problems and two different sets of solutions. Any recommendation built on this work should say plainly which of the two problems it addresses, because the evidence here is about as clear as applied social-science evidence gets that they are not the same thing.

<br>

# Multivariate Confirmation: Side-by-Side Comparison of the Two Disconnection Mechanisms

**Non-College Graduates, Southwest Minnesota (2011–2019 Cohorts)**

**Variable Screening Table** (final-model results, N = 8,707)

| Variable | Bivariate Expectation | Confirmed for Non-Sustained? | Confirmed for No-Record? |
|------------------|------------------|------------------|------------------|
| Gender | Non-sustained mechanism (primarily) | **Yes** (OR = 1.47, harmful) | **No** — null in final model (OR = 1.07, p = .217) |
| RaceEthnicity — American Indian | No-record mechanism (primarily) | **No** — null in every model | **Yes** (OR = 1.74, harmful) |
| RaceEthnicity — Black | Non-sustained mechanism (primarily) | **Yes**, strongly and strengthening (OR = 1.84, harmful) | **No** — null in every model (OR = 0.88, p = .485) |
| RaceEthnicity — Asian/PI | No-record mechanism (primarily) | **No** — null throughout | **Yes** (OR = 1.38, harmful) |
| RaceEthnicity — Hispanic | No-record mechanism (primarily) | **No** — null throughout (OR = 1.08, p = .485) | **Yes** (OR = 1.51, harmful) |
| economic.status | Non-sustained mechanism (primarily) | **Yes** (OR = 1.30, harmful) | **Yes, opposite direction** (OR = 0.66, protective) |
| SpecialEdStatus | Non-sustained mechanism | **Yes, strongly, and strengthens across blocks** (OR = 1.53, harmful) | **No** — null from Model 2 onward (OR = 0.95, p = .432) |
| english.learner | No-record mechanism (weakening) | **No** — null in every model | **Marginal only** — significant in Models 1–2, fades by final model (OR = 1.15, p = .107) |
| Dem_Desc — Town/rural mix | Both, opposite directions | **Yes** (OR = 1.25, harmful) | **Yes, opposite direction** (OR = 0.54, protective) |
| Dem_Desc — Urban/town/rural mix | Weak, unclear | **No** | **Marginal** (OR = 0.83, p = .095) |
| edr — EDR 8 | No-record mechanism (primarily) | **Yes, protective** (OR = 0.81, protective) | **Marginal, harmful** (OR = 1.14, p = .076) — attenuated once accomplishments/attended.ps entered |
| edr — EDR 6E | No-record mechanism | **No** | **Yes** (OR = 0.66, protective) |
| avg.wages.pct.state | No-record mechanism | **Marginal** (OR = 2.72, p = .049) | **Yes** (OR = 0.31, protective) |
| avg.unemp.rate | No-record mechanism | **No** (p = .953) | **Yes** (OR = 0.92, protective) |
| pseo.participant | No-record mechanism | **No** (p = .458) | **Yes** (OR = 0.76, protective) |
| took.ACT | No-record mechanism, dramatically | **No** (p = .515) | **Yes, dramatically** (OR = 0.28, protective) — the largest effect in either sequence |
| ACTCompositeScore.filled | No-record mechanism | **Marginal only** (OR = 1.02, p = .087) | **Yes** (OR = 1.04, harmful) |
| ap.exam | No relationship (bivariate null) | **No** (p = .206) | **No** (p = .765) |
| sat.taken | No relationship (small subgroup) | **No** (p = .889) | **No** (p = .269) |
| cte.achievement | No-record mechanism, strongly | **No**, both categories null (p = .844, .488) | **Yes**, both categories (OR = 0.69 / 0.64, protective) |
| total.cte.courses.taken | No-record mechanism, strongly | **No** (p = .636) | **Yes** (OR = 0.92, protective) |
| attended.ps | Both, roughly balanced | **Yes** (OR = 0.64, protective) | **Yes, similar strength** (OR = 0.78, protective) |

**Description of the Analysis and Findings**

This analysis tested, via parallel four-block logistic regression sequences (demographics → high school characteristics → high school accomplishments → postsecondary attempt), whether the two disconnection mechanisms identified in the bivariate work — persistent non-sustained employment and persistent no-record status — remain genuinely distinct once all variables are considered simultaneously. Both models were fit on the same analytic population (N = 8,707 non-college graduates, 2011–2019 cohorts), using the same 16-variable candidate set with identical block structure.

The results confirm the mechanism split with unusual cleanliness. Unlike a comparable analysis in another region, where several variables (Hispanic ethnicity, economic status, SAT-taking) showed real, opposite-direction effects on *both* mechanisms, this Southwest Minnesota population shows almost no genuine dual-mechanism variables outside of `attended.ps` and, to a lesser degree, `economic.status` and community type. Nearly every other variable sorted into a single dominant mechanism with the other side reaching a clean, non-significant null — a sharper split than the comparison region's more overlapping pattern.

**Economic status and community type (town/rural mix) are the two clearest "opposite direction" findings in this analysis.** Economic status is harmful for non-sustained employment (OR = 1.30) and protective for no-record status (OR = 0.66) — economically disadvantaged graduates are more likely to cycle through unstable work, but less likely to disappear from Minnesota's employment record altogether, most plausibly because economic constraint limits the option of leaving the local labor market or state entirely. Town/rural mix community background shows the identical reversal in miniature: a modest risk factor for job instability (OR = 1.25) but a strong protective factor against no-record status (OR = 0.54). Combining these two outcomes into a single blended measure, as the original bivariate disconnection framing did, would dilute or wash out both relationships rather than merely averaging them.

**Gender and special education status are the cleanest single-mechanism confirmations in this analysis.** Gender is significant and harmful for non-sustained employment in every model and fully null for no-record status by the final specification. Special education status shows the identical pattern, and more emphatically — strengthening across every block of the non-sustained sequence while never approaching significance for no-record status past the demographics-only stage. This directly corrects the analysis's own introductory hypothesis, which characterized special education status as a no-record-mechanism variable; the evidence, now confirmed under full multivariate control twice over, says the opposite.

**American Indian, Asian/PI, Black, and Hispanic graduates' disparities — treated as a single "RaceEthnicity" finding at the bivariate stage — resolve into genuinely distinct, almost entirely non-overlapping group-level stories.** American Indian, Asian/PI, and Hispanic graduates all show real, significant risk of persistent no-record status and no relationship whatsoever with non-sustained employment at any model stage. Black graduates show the complete opposite concentration: the largest and most durable effect found in either sequence for non-sustained employment (OR = 1.84, strengthening across blocks) and no relationship with no-record status at all. Treating "racial disparity in disconnection" as one undifferentiated story, as the original combined bivariate measure did, would have entirely missed that these four groups face two structurally different versions of workforce disconnection.

**English learner status shows a genuine mediation effect not present for most other mechanism-specific variables.** It is significant for no-record status through the demographics and high-school-characteristics stages, but fades to non-significance once accomplishments are added — suggesting its real, bivariate-confirmed no-record risk substantially runs through differential accomplishment access (most plausibly the documented English-learner ACT-participation gap) rather than persisting as a fully independent effect once the complete model is specified. It shows no relationship with non-sustained employment at any stage.

**took.ACT and ACTCompositeScore.filled deliver the single cleanest mechanism-specific confirmation in the entire analysis.** took.ACT shows no relationship with non-sustained employment (p = .515) but the single strongest protective effect found in either sequence against no-record status (OR = 0.28). ACTCompositeScore.filled is only marginal for non-sustained employment but clearly significant and harmful for no-record status — replicating the "engagement helps, achievement hurts" tension documented throughout this project's completer-focused work, now localized specifically to the record-disappearance mechanism among non-completers as well.

**CTE engagement is, in this region, a clean no-record-only finding across every specification — not the categorical/continuous split seen elsewhere.** Both CTE achievement categories and total CTE courses taken are large, significant, and protective against no-record status, and every one of these same terms is fully null against non-sustained employment. Unlike a comparable regional analysis, where CTE achievement showed some balanced protection across both mechanisms, this population's CTE benefit is concentrated entirely in keeping graduates visible in Minnesota's wage data.

**`attended.ps` is the sole variable confirmed as genuinely dual-mechanism, and its effect is close to symmetric.** It is strongly protective for non-sustained employment (OR = 0.64) and similarly protective for no-record status (OR = 0.78) — the only variable in either sequence to show a real, significant, independent effect on both outcomes, exactly matching its unique bivariate profile and standing as the clearest actionable, dual-purpose lever identified anywhere in this non-completer analysis.

**Local wages, local unemployment, and EDR all show genuine mechanism-specific regional signatures.** Local average wages and local unemployment rate are both null for non-sustained employment but significant for no-record status (wages protective, unemployment protective — both consistent with the interpretation that economic constraint functions as a tether keeping graduates visible in the record). EDR splits cleanly by mechanism and by region: EDR 8 is significantly protective for job instability but only marginally harmful for no-record status once fully controlled; EDR 6E shows no non-sustained association but is significantly protective for no-record status. Different regional and economic signals matter for different mechanisms — not a single, unified "regional disadvantage" story.

**The Story the Data Tells**

Coming into this analysis, the working hypothesis — built from the bivariate stage — was that "disconnection" among non-college graduates might really be two problems wearing one label. This multivariate confirmation supports that hypothesis about as cleanly as applied social-science evidence gets. Nearly every variable tested sorted into a single dominant mechanism, with the other side reaching a genuine, stable, non-significant null rather than a partial or ambiguous signal. Only `attended.ps` shows a real, independent effect on both outcomes, and only economic status and community type show the more complicated "opposite direction" pattern seen more broadly in comparable regional analyses.

If your risk factor is gender or special education status, the data says only one mechanism applies to you at all — you face real, elevated risk of cycling through unstable employment, and no elevated risk of disappearing from Minnesota's workforce record. If your risk factor is being American Indian, Asian/PI, or Hispanic, the reverse is true entirely: real, elevated risk of vanishing from the record, and no elevated risk of instability. If you are Black, your risk is concentrated in instability, the mirror image of the three groups just named. If you are economically disadvantaged, both mechanisms apply to you, but in opposite ways — a genuinely elevated risk of unstable work, alongside a genuinely reduced risk of disappearing from the system altogether, most plausibly because economic necessity keeps you working locally even when that work is unstable.

This has a direct, practical implication for anyone designing an intervention around this population. Programs built around ACT participation or CTE achievement would show large, measurable results against record-disappearance, with essentially no effect on job stability. Programs built around special-education support services or targeted at gender-specific barriers would move the needle on job stability specifically, with no measurable effect on the record-disappearance problem. Programs built around postsecondary attendance — even incomplete attendance — would help meaningfully with both, roughly evenly, making `attended.ps` the single most broadly useful lever identified anywhere in this two-part confirmatory analysis. And any intervention built around a single, undifferentiated theory of "workforce disconnection" — the kind implied by the original combined 50.4%-persistently-disconnected headline figure — risks helping one population while doing nothing measurable for another, particularly for the demographic groups (American Indian, Asian/PI, Hispanic, and Black graduates) whose disparities, treated as one undifferentiated racial gap, actually describe two entirely separate phenomena requiring two entirely separate responses.
